---
title: "Section 4: Gapminder"
output: html_notebook
---

# Overview

In Section 4, you will look at a case study involving data from the Gapminder Foundation about trends in world health and economics.

After completing Section 4, you will:

* understand how Hans Rosling and the Gapminder Foundation use effective data visualization to convey data-based trends.  
* be able to apply the ggplot2 techniques from the previous section to answer questions using data.  
* understand how fixed scales across plots can ease comparisons.
be able to modify graphs to improve data visualization.  

# Introduction to Gapminder

## Case Study: Trends in World Health and Economics

**More about Gapminder**

The original Gapminder TED talks are available and we encourage you to watch them.

* The Best Stats You've Ever Seen  
* New Insights on Poverty  

You can also find more information and raw data (in addition to what we analyze in class) at https://www.gapminder.org/.

**Key points**

* Data visualization can be used to dispel common myths and educate the public and contradict sensationalist or outdated claims and stories.  
* We will use real data to answer the following questions about world health and economics:  
  + Is it still fair to consider the world as divided into the West and the developing world?  
  + Has income inequality across countries worsened over the last 40 years?  


## Gapminder Dataset

**Key points**

* A selection of world health and economics statistics from the Gapminder project can be found in the dslabs package as data(gapminder).  
* Most people have misconceptions about world health and economics, which can be addressed by considering real data.  

**Code**

```{r}
# load and inspect gapminder data
library(dslabs)
data(gapminder)
head(gapminder)

# compare infant mortality in Sri Lanka and Turkey
gapminder %>%
    filter(year == 2015 & country %in% c("Sri Lanka", "Turkey")) %>%
    select(country, infant_mortality)
```


## Life Expectancy and Fertility Rates

**Key points**

* A prevalent worldview is that the world is divided into two groups of countries:  
  + Western world: high life expectancy, low fertility rate  
  + Developing world: lower life expectancy, higher fertility rate  
* Gapminder data can be used to evaluate the validity of this view.  
* A scatterplot of life expectancy versus fertility rate in 1962 suggests that this viewpoint was grounded in reality 50 years ago. Is it still the case today?  

**Code**

```{r}
# basic scatterplot of life expectancy versus fertility
ds_theme_set()    # set plot theme
filter(gapminder, year == 1962) %>%
    ggplot(aes(fertility, life_expectancy)) +
    geom_point()

# add color as continent
filter(gapminder, year == 1962) %>%
    ggplot(aes(fertility, life_expectancy, color = continent)) +
    geom_point()
```

