---
title: "Section 4: Gapminder"
output: html_notebook
---

# Overview

In Section 4, you will look at a case study involving data from the Gapminder Foundation about trends in world health and economics.

After completing Section 4, you will:

* understand how Hans Rosling and the Gapminder Foundation use effective data visualization to convey data-based trends.  
* be able to apply the ggplot2 techniques from the previous section to answer questions using data.  
* understand how fixed scales across plots can ease comparisons.
be able to modify graphs to improve data visualization.  


# 4.2 Using the Gapminder Dataset Assessment


## 4.2.1 Faceting

**Key points**

* Faceting makes multiple side-by-side plots stratified by some variable. This is a way to ease comparisons.  
* The facet_grid function allows faceting by up to two variables, with rows faceted by one variable and columns faceted by the other variable. To facet by only one variable, use the dot operator as the other variable.  
* The facet_wrap function facets by one variable and automatically wraps the series of plots so they have readable dimensions.  
* Faceting keeps the axes fixed across all plots, easing comparisons between plots.  
* The data suggest that the developing versus Western world view no longer makes sense in 2012.

**Code**

```{r}
# facet by continent and year
filter(gapminder, year %in% c(1962, 2012)) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point() +
    facet_grid(continent ~ year)

# facet by year only
filter(gapminder, year %in% c(1962, 2012)) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point() +
    facet_grid(. ~ year)

# facet by year, plots wrapped onto multiple rows
years <- c(1962, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
gapminder %>%
    filter(year %in% years & continent %in% continents) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point() +
    facet_wrap(~year)
```



## 4.2.2 Time Series Plots

**Key points**

* Time series plots have time on the x-axis and a variable of interest on the y-axis.  
* The geom_line geometry connects adjacent data points to form a continuous line. A line plot is appropriate when points are regularly spaced, densely packed and from a single data series.  
*z You can plot multiple lines on the same graph. Remember to group or color by a variable so that the lines are plotted independently.  
* Labeling is usually preferred over legends. However, legends are easier to make and appear by default. Add a label with geom_text, specifying the coordinates where the label should appear on the graph.  

**Code: Single time series**

```{r}
# scatterplot of US fertility by year
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_point()

# line plot of US fertility by year
gapminder %>%
    filter(country == "United States") %>%
    ggplot(aes(year, fertility)) +
    geom_line()
```

**Code: Multiple time series**

```{r}
# line plot fertility time series for two countries- only one line (incorrect)
countries <- c("South Korea", "Germany")
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility)) +
    geom_line()

# line plot fertility time series for two countries - one line per country
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility, group = country)) +
    geom_line()

# fertility time series for two countries - lines colored by country
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, fertility, col = country)) +
    geom_line()
```

**Code: Adding text labels to a plot**

```{r}
# life expectancy time series - lines colored by country and labeled, no legend
labels <- data.frame(country = countries, x = c(1975, 1965), y = c(60, 72))
gapminder %>% filter(country %in% countries) %>%
    ggplot(aes(year, life_expectancy, col = country)) +
    geom_line() +
    geom_text(data = labels, aes(x, y, label = country), size = 5) +
    theme(legend.position = "none")
```



## 4.2.3 Transformations

**Key points**

* We use GDP data to compute income in US dollars per day, adjusted for inflation.  
* Log transformations convert multiplicative changes into additive changes.  
* Common transformations are the log base 2 transformation and the log base 10 transformation. The choice of base depends on the range of the data. The natural log is not recommended for visualization because it is difficult to interpret.  
* The mode of a distribution is the value with the highest frequency. The mode of a normal distribution is the average. A distribution can have multiple local modes.  
* There are two ways to use log transformations in plots: transform the data before plotting or transform the axes of the plot. Log scales have the advantage of showing the original values as axis labels, while log transformed values ease interpretation of intermediate values between labels.  
* Scale the x-axis using scale_x_continuous or scale_x_log10 layers in ggplot2. Similar functions exist for the y-axis.  
* In 1970, income distribution is bimodal, consistent with the dichotomous Western versus developing worldview.  

**Code**

```{r}
# add dollars per day variable
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)
    
# histogram of dollars per day
past_year <- 1970
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black")
    
# repeat histogram with log2 scaled data
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(log2(dollars_per_day))) +
    geom_histogram(binwidth = 1, color = "black")
    
# repeat histogram with log2 scaled x-axis
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2")
```



## 4.2.4 Stratify and Boxplot

**Key points**

* Make boxplots stratified by a categorical variable using the geom_boxplot geometry.  
* Rotate axis labels by changing the theme through element_text. You can change the angle and justification of the text labels.  
* Consider ordering your factors by a meaningful value with the reorder function, which changes the order of factor levels based on a related numeric vector. This is a way to ease comparisons.  
* Show the data by adding data points to the boxplot with a geom_point layer. This adds information beyond the five-number summary to your plot, but too many data points it can obfuscate your message.  

**Code: Boxplot of GDP by region**

```{r}
# add dollars per day variable
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)

# number of regions
length(levels(gapminder$region))

# boxplot of GDP by region in 1970
past_year <- 1970
p <- gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    ggplot(aes(region, dollars_per_day))
p + geom_boxplot()

# rotate names on x-axis
p + geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Code: The reorder function**

```{r}
# by default, factor order is alphabetical
fac <- factor(c("Asia", "Asia", "West", "West", "West"))
levels(fac)

# reorder factor by the category means
value <- c(10, 11, 12, 6, 4)
fac <- reorder(fac, value, FUN = mean)
levels(fac)
```

**Code: Enhanced boxplot ordered by median income, scaled, and showing data**

```{r}
# reorder by median income and color by continent
p <- gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%    # reorder
    ggplot(aes(region, dollars_per_day, fill = continent)) +    # color by continent
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("")
p

# log2 scale y-axis
p + scale_y_continuous(trans = "log2")

# add data points
p + scale_y_continuous(trans = "log2") + geom_point(show.legend = FALSE)
```



## 4.2.5 Comparing Distributions

**Key points**

* Use intersect to find the overlap between two vectors.  
* To make boxplots where grouped variables are adjacaent, color the boxplot by a factor instead of faceting by that factor. This is a way to ease comparisons.  
* The data suggest that the income gap between rich and poor countries has narrowed, not expanded.  

**Code: Histogram of income in West versus developing world, 1970 and 2010**

```{r}
# add dollars per day variable and define past year
gapminder <- gapminder %>%
    mutate(dollars_per_day = gdp/population/365)
past_year <- 1970

# define Western countries
west <- c("Western Europe", "Northern Europe", "Southern Europe", "Northern America", "Australia and New Zealand")

# facet by West vs devloping
gapminder %>%
    filter(year == past_year & !is.na(gdp)) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(. ~ group)

# facet by West/developing and year
present_year <- 2010
gapminder %>%
    filter(year %in% c(past_year, present_year) & !is.na(gdp)) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(year ~ group)
```

**Code: Income distribution of West versus developing world, only countries with data**

```{r}
# define countries that have data available in both years
country_list_1 <- gapminder %>%
    filter(year == past_year & !is.na(dollars_per_day)) %>% .$country
country_list_2 <- gapminder %>%
    filter(year == present_year & !is.na(dollars_per_day)) %>% .$country
country_list <- intersect(country_list_1, country_list_2)

# make histogram including only countries with data available in both years
gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%    # keep only selected countries
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day)) +
    geom_histogram(binwidth = 1, color = "black") +
    scale_x_continuous(trans = "log2") +
    facet_grid(year ~ group)
```

**Code: Boxplots of income in West versus developing world, 1970 and 2010**

```{r}
p <- gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%
    ggplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("") + scale_y_continuous(trans = "log2")
p + geom_boxplot(aes(region, dollars_per_day, fill = continent)) +
    facet_grid(year ~ .)

# arrange matching boxplots next to each other, colored by year
p + geom_boxplot(aes(region, dollars_per_day, fill = factor(year)))
```



## 4.2.6 Density Plots

**Key points**

* Change the y-axis of density plots to variable counts using ..count.. as the y argument.  
* The case_when function defines a factor whose levels are defined by a variety of logical operations to group data.  
* Plot stacked density plots using position="stack".  
* Define a weight aesthetic mapping to change the relative weights of density plots - for example, this allows weighting of plots by population rather than number of countries.

**Code: Faceted smooth density plots**

```{r}
# see the code below the previous video for variable definitions

# smooth density plots - area under each curve adds to 1
gapminder %>%
    filter(year == past_year & country %in% country_list) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>% group_by(group) %>%
    summarize(n = n()) %>% knitr::kable()

# smooth density plots - variable counts on y-axis
p <- gapminder %>%
    filter(year == past_year & country %in% country_list) %>%
    mutate(group = ifelse(region %in% west, "West", "Developing")) %>%
    ggplot(aes(dollars_per_day, y = ..count.., fill = group)) +
    scale_x_continuous(trans = "log2")
p + geom_density(alpha = 0.2, bw = 0.75) + facet_grid(year ~ .)
```

**Code: Add new region groups with case_when**

```{r}
# add group as a factor, grouping regions
gapminder <- gapminder %>%
    mutate(group = case_when(
        .$region %in% west ~ "West",
        .$region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
        .$region %in% c("Caribbean", "Central America", "South America") ~ "Latin America",
        .$continent == "Africa" & .$region != "Northern Africa" ~ "Sub-Saharan Africa",
        TRUE ~ "Others"))

# reorder factor levels
gapminder <- gapminder %>%
    mutate(group = factor(group, levels = c("Others", "Latin America", "East Asia", "Sub-Saharan Africa", "West")))
```

**Code: Stacked density plot**

```{r}
# note you must redefine p with the new gapminder object first
p <- gapminder %>%
  filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    ggplot(aes(dollars_per_day, fill = group)) +
    scale_x_continuous(trans = "log2")

# stacked density plot
p + geom_density(alpha = 0.2, bw = 0.75, position = "stack") +
    facet_grid(year ~ .)
```

**Code: Weighted stacked density plot**

```{r}
# weighted stacked density plot
gapminder %>%
    filter(year %in% c(past_year, present_year) & country %in% country_list) %>%
    group_by(year) %>%
    mutate(weight = population/sum(population*2)) %>%
    ungroup() %>%
    ggplot(aes(dollars_per_day, fill = group, weight = weight)) +
    scale_x_continuous(trans = "log2") +
    geom_density(alpha = 0.2, bw = 0.75, position = "stack") + facet_grid(year ~ .)
```



## 4.2.7 Ecological Fallacy

**Key points**

* The breaks argument allows us to set the location of the axis labels and tick marks.
* The *logistic* or *logit transformation* is defined as $f(p) = log\frac{p}{1-p}$, or the log of odds. This scale is useful for highlighting differences near 0 or near 1 and converts fold changes into constant increases.  
* The *ecological fallacy* is assuming that conclusions made from the average of a group apply to all members of that group.  

**Code**

```{r}
# define gapminder
library(tidyverse)
library(dslabs)
data(gapminder)

# add additional cases
gapminder <- gapminder %>%
    mutate(group = case_when(
        .$region %in% west ~ "The West",
        .$region %in% "Northern Africa" ~ "Northern Africa",
        .$region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
        .$region == "Southern Asia" ~ "Southern Asia",
        .$region %in% c("Central America", "South America", "Caribbean") ~ "Latin America",
        .$continent == "Africa" & .$region != "Northern Africa" ~ "Sub-Saharan Africa",
        .$region %in% c("Melanesia", "Micronesia", "Polynesia") ~ "Pacific Islands"))

# define a data frame with group average income and average infant survival rate
surv_income <- gapminder %>%
    filter(year %in% present_year & !is.na(gdp) & !is.na(infant_mortality) & !is.na(group)) %>%
    group_by(group) %>%
    summarize(income = sum(gdp)/sum(population)/365,
                        infant_survival_rate = 1 - sum(infant_mortality/1000*population)/sum(population))
surv_income %>% arrange(income)

# plot infant survival versus income, with transformed axes
surv_income %>% ggplot(aes(income, infant_survival_rate, label = group, color = group)) +
    scale_x_continuous(trans = "log2", limit = c(0.25, 150)) +
    scale_y_continuous(trans = "logit", limit = c(0.875, .9981),
                                       breaks = c(.85, .90, .95, .99, .995, .998)) +
    geom_label(size = 3, show.legend = FALSE) 
```

