---
title: "Section 3: Random Variables, Sampling Models, and the Central Limit Theorem"
output: html_notebook
---

# 3.0 Overview


Section 3 introduces you to Random Variables, Sampling Models, and the Central Limit Theorem.

Section 3 is divided into two parts:

* Random Variables and Sampling Models
* The Central Limit Theorem.

After completing Section 3, you will:

* understand what random variables are, how to generate them, and the correct mathematical notation to use with them.
* be able to use sampling models to estimate characteristics of a larger population.
* be able to explain the difference between a distribution and a probability distribution.
* understand the Central Limit Theorem and the law of large numbers.

There are 2 assignments that use the DataCamp platform for you to practice your coding skills as well as a set of questions on the edX platform at the end of Section 3.  

This section corresponds to the following section of the course textbook.  

We encourage you to use R to interactively test out your answers and further your learning.


# 3.1 Random Variables and Sampling Models

## 3.1.1 Random Variables

**Key points**

* Random variables are numeric outcomes resulting from random processes.
* Statistical inference offers a framework for quantifying uncertainty due to randomness.

**Code: Modeling a random variable**

```{r}
# define random variable x to be 1 if blue, 0 otherwise
beads <- rep(c("red", "blue"), times = c(2, 3))
x <- ifelse(sample(beads, 1) == "blue", 1, 0)

# demonstrate that the random variable is different every time
ifelse(sample(beads, 1) == "blue", 1, 0)
ifelse(sample(beads, 1) == "blue", 1, 0)
ifelse(sample(beads, 1) == "blue", 1, 0)
```

## 3.1.2 Sampling Models

**Key points**

* A sampling model models the random behavior of a process as the sampling of draws from an urn.
* The probability distribution of a random variable is the probability of the observed value falling in any given interval.
* We can define a CDF  ????(????)=Pr(???????????)  to answer questions related to the probability of S being in any interval.
* The average of many draws of a random variable is called its expected value.
* The standard deviation of many draws of a random variable is called its standard error.

**Monte Carlo simulation: Chance of casino losing money on roulette**

We build a sampling model for the random variable  ????  that represents the casino's total winnings. 

```{r}
# sampling model 1: define urn, then sample
color <- rep(c("Black", "Red", "Green"), c(18, 18, 2)) # define the urn for the sampling model
n <- 1000
X <- sample(ifelse(color == "Red", -1, 1), n, replace = TRUE)    # 1000 draws from urn, -1 if red, else +1
X[1:10]    # first 10 outcomes

# sampling model 2: define urn inside sample function by noting probabilities
x <- sample(c(-1, 1), n, replace = TRUE, prob = c(9/19, 10/19))    # 1000 independent draws
S <- sum(x)    # total winnings = sum of draws
S
```

We use the sampling model to run a Monte Carlo simulation and use the results to estimate the probability of the casino losing money.

```{r}
n <- 1000    # number of roulette players
B <- 10000    # number of Monte Carlo experiments
S <- replicate(B, {
    X <- sample(c(-1,1), n, replace = TRUE, prob = c(9/19, 10/19))    # simulate 1000 roulette spins
    sum(X)    # determine total profit
})

mean(S < 0)    # probability of the casino losing money
```

We can plot a histogram of the observed values of S as well as the normal density curve based on the mean and standard deviation of S.

```{r}
library(tidyverse)

s <- seq(min(S), max(S), length = 100)    # sequence of 100 values across range of S
normal_density <- data.frame(s = s, f = dnorm(s, mean(S), sd(S))) # generate normal density for S
data.frame (S = S) %>%    # make data frame of S for histogram
    ggplot(aes(S, stat(density))) +
    geom_histogram(color = "black", binwidth = 10) +
    ylab("Probability") +
    geom_line(data = normal_density, mapping = aes(s, f), color = "blue")
```


## 3.1.3 Distributions versus Probability Distributions

* A random variable *X* has a probability distribution function *F(a)* that defines $Pr(X \le a)$ over all values of *a*.
* Any list of numbers has a distribution. The probability distribution function of a random variable is defined mathematically and does not depend on a list of numbers.
* The results of a Monte Carlo simulation with a large enough number of observations will approximate the probability distribution of *X*.
* If a random variable is defined as draws from an urn:
  + The probability distribution function of the random variable is defined as the distribution of the list of values in the urn.
  + The expected value of the random variable is the average of values in the urn.
  + The standard error of one draw of the random variable is the standard deviation of the values of the urn.


## 3.1.4 Notation for Random Variables

**Key points**

* Capital letters denote random variables *(X)* and lowercase letters denote observed values *(x)*.
* In the notation $Pr(X=x)$, we are asking how frequently the random variable *X* is equal to the value x . For example, if *x=6*, this statement becomes $Pr(X=6)$.


## 3.1.5 Central Limit Theorem (CLT)

**Key points**

* The Central Limit Theorem (CLT) says that the distribution of the sum of a random variable is approximated by a normal distribution.
* The expected value of a random variable, $E[X] = \mu$, is the average of the values in the urn. This represents the expectation of one draw. 
* The standard error of one draw of a random variable is the standard deviation of the values in the urn.
* The expected value of the sum of draws is the number of draws times the expected value of the random variable. 
* The standard error of the sum of independent draws of a random variable is the square root of the number of draws times the standard deviation of the urn. 

**Equations**  
These equations apply to the case where there are only two outcomes, *a* and *b* with proportions *p* and *1-p* respectively. The general principles above also apply to random variables with more than two outcomes.

*Expected value of a random variable: *
$$ap + b(1-p)$$
*Expected value of the sum of n draws of a random variable: *
$$n*(ap + b(1-p))$$
*Standard deviation of an urn with two values: *
$$|b-a| \sqrt{p(1-p)}$$

*Standard error of the sum of n draws of a random variable:*
$$\sqrt{n}*|b-a| \sqrt{p(1-p)}$$

## 3.1.6 Assessment

### Exercise 1. American Roulette probabilities
An American roulette wheel has 18 red, 18 black, and 2 green pockets. Each red and black pocket is associated with a number from 1 to 36. The two remaining green slots feature "0" and "00". Players place bets on which pocket they think a ball will land in after the wheel is spun. Players can bet on a specific number (0, 00, 1-36) or color (red, black, or green).

What are the chances that the ball lands in a green pocket?

*Instructions*

* Define a variable p_green as the probability of the ball landing in a green pocket.
* Print the value of p_green.

*Code*
```{r}
# The variables `green`, `black`, and `red` contain the number of pockets for each color
green <- 2
black <- 18
red <- 18

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- 2/38

# Print the variable `p_green` to the console
p_green
```

### Exercise 2. American Roulette payout
In American roulette, the payout for winning on green is $17. This means that if you bet $1 and it lands on green, you get $17 as a prize.

Create a model to predict your winnings from betting on green one time.

*Instructions*

* Use the sample function return a random value from a specified range of values.
* Use the prob = argument in the sample function to specify a vector of probabilities for returning each of the values contained in the vector of values being sampled.
* Take a single sample (n = 1).

*Code*
```{r}
# Use the `set.seed` function to make sure your answer matches the expected result after random sampling.
set.seed(1)

# The variables 'green', 'black', and 'red' contain the number of pockets for each color
green <- 2
black <- 18
red <- 18

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- green / (green+black+red)

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Create a model to predict the random variable `X`, your winnings from betting on green. Sample one time.
n <- 1
X <- sample(c(17,-1,-1), n, replace=TRUE, prob=c(p_green, p_not_green/2, p_not_green/2))

# Print the value of `X` to the console
X

```

### Exercise 3. American Roulette expected value
In American roulette, the payout for winning on green is $17. This means that if you bet $1 and it lands on green, you get $17 as a prize.In the previous exercise, you created a model to predict your winnings from betting on green.

Now, compute the expected value of X, the random variable you generated previously.

*Instructions*

* Using the chances of winning $17 (p_green) and the chances of losing $1 (p_not_green), calculate the expected outcome of a bet that the ball will land in a green pocket.

*Code*
```{r}
# The variables 'green', 'black', and 'red' contain the number of pockets for each color
green <- 2
black <- 18
red <- 18

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- green / (green+black+red)

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Calculate the expected outcome if you win $17 if the ball lands on green and you lose $1 if the ball doesn't land on green
p_green * 17 + p_not_green *(-1)

```

### Exercise 4. American Roulette standard error
The standard error of a random variable X tells us the difference between a random variable and its expected value. You calculated a random variable X in exercise 2 and the expected value of that random variable in exercise 3.

Now, compute the standard error of that random variable, which represents a single outcome after one spin of the roulette wheel.

*Instructions*

* Compute the standard error of the random variable you generated in exercise 2, or the outcome of any one spin of the roulette wheel.
* Recall that the payout for winning on green is $17 for a $1 bet.

*Code*
```{r}
# The variables 'green', 'black', and 'red' contain the number of pockets for each color
green <- 2
black <- 18
red <- 18

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- green / (green+black+red)

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Compute the standard error of the random variable
abs(17 - (-1)) * sqrt(p_green * p_not_green)
```

### Exercise 5. American Roulette sum of winnings
You modeled the outcome of a single spin of the roulette wheel, X, in exercise 2.
Now create a random variable S that sums your winnings after betting on green 1,000 times.

*Instructions*

* Use set.seed to make sure the result of your random operation matches the expected answer for this problem.
* Specify the number of times you want to sample from the possible outcomes.
* Use the sample function to return a random value from a vector of possible values.
* Be sure to assign a probability to each outcome and to indicate that you are sampling with replacement.

*Code*
```{r}
# The variables 'green', 'black', and 'red' contain the number of pockets for each color
green <- 2
black <- 18
red <- 18

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- green / (green+black+red)

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Use the `set.seed` function to make sure your answer matches the expected result after random sampling
set.seed(1)

# Define the number of bets using the variable 'n'
n <- 1000

# Create a vector called 'X' that contains the outcomes of 1000 samples
X <- sample(c(17,-1,-1), n, replace=TRUE, prob=c(p_green,p_not_green/2,p_not_green/2))

# Assign the sum of all 1000 outcomes to the variable 'S'
S <- sum(X)

# Print the value of 'S' to the console
S
```

### Exercise 6. American Roulette winnings expected valu
In the previous exercise, you generated a vector of random outcomes, S, after betting on green 1,000 times.
What is the expected value of S?

*Instructions*

* Using the chances of winning $17 (p_green) and the chances of losing $1 (p_not_green), calculate the expected outcome of a bet that the ball will land in a green pocket over 1,000 bets.

*Code*
```{r}
# The variables 'green', 'black', and 'red' contain the number of pockets for each color
green <- 2
black <- 18
red <- 18

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- green / (green+black+red)

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Define the number of bets using the variable 'n'
n <- 1000

# Calculate the expected outcome of 1,000 spins if you win $17 when the ball lands on green and you lose $1 when the ball doesn't land on green
n * (p_green * 17 + p_not_green * (-1))
```

### Exercise 7. American Roulette winnings expected value
You generated the expected value of S, the outcomes of 1,000 bets that the ball lands in the green pocket, in the previous exercise.
What is the standard error of S?

*Instructions*

* Compute the standard error of the random variable you generated in exercise 5, or the outcomes of 1,000 spins of the roulette wheel.

*Code*
```{r}
# The variables 'green', 'black', and 'red' contain the number of pockets for each color
green <- 2
black <- 18
red <- 18

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- green / (green+black+red)

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Define the number of bets using the variable 'n'
n <- 1000

# Compute the standard error of the sum of 1,000 outcomes
n * abs(17 - (-1)) * sqrt(p_green * p_not_green)
```
