---
title: "Section 3: Random Variables, Sampling Models, and the Central Limit Theorem"
output: html_notebook
---

# 3.0 Overview


Section 3 introduces you to Random Variables, Sampling Models, and the Central Limit Theorem.

Section 3 is divided into two parts:

* Random Variables and Sampling Models
* The Central Limit Theorem.

After completing Section 3, you will:

* understand what random variables are, how to generate them, and the correct mathematical notation to use with them.
* be able to use sampling models to estimate characteristics of a larger population.
* be able to explain the difference between a distribution and a probability distribution.
* understand the Central Limit Theorem and the law of large numbers.

There are 2 assignments that use the DataCamp platform for you to practice your coding skills as well as a set of questions on the edX platform at the end of Section 3.  

This section corresponds to the following section of the course textbook.  

We encourage you to use R to interactively test out your answers and further your learning.

```{r}
library(gtools)
library(tidyverse)
options(digits = 3)    # report 3 significant digits
```

# 3.2 The Central Limit Theorem Continued

## 3.2.1 Averages and Proportions

**Key points**

* Random variable times a constant
  + *The expected value of a random variable multiplied by a constant* is that constant times its original expected value:
$$E[aX] = a\mu$$

  + *The standard error of a random variable multiplied by a constant* is that constant times its original standard error:
$$E[aX] = a\sigma$$

* Average of multiple draws of a random variable
  + *The expected value of average of multiple draws* from an urn is the expected value of the urn ($\mu$)

  + *The standard deviation of the average of multiple draws* from an urn is the standard deviation of the urn divided by the square root of the number of draws ($sigma/\sqrt{n}$)

* The sum of multiple draws of a random variable
  + *The expected value of the sum of n draws* of random variable is *n* times its original expected value:
$$E[nX] = n\mu$$

  + *The standard error of the sum of n draws of random variable* is $sqrt{n}$ times its original standard error:
$$E[nX] = \sqrt{n}\sigma$$

* The sum of multiple different random variables
  + *The expected value of the sum of different random variables* is the sum of the individual expected values for each random variable:
$$E[X_1+X_2+...+X_n] = \mu_1+\mu_2+...+\mu_n$$

  + *The standard error of the sum of different random variables* is the square root of the sum of squares of the individual standard errors:
$$E[X_1+X_2+...+X_n] = \sqrt{\sigma_1^2+\sigma_2^2+...+\sigma_n^2}$$

* Transformation of random variables
  + If *X* is a normally distributed random variable and *a* and *b* are non-random constants, then *aX+b* is also a normally distributed random variable.


## 3.2.2 Law of Large Numbers

**Key points**

* The law of large numbers states that as *n* increases, the standard error of the average of a random variable decreases. In other words, when *n* is large, the average of the draws converges to the average of the urn.
* The law of large numbers is also known as the law of averages.
* The law of averages only applies when *n* is very large and events are independent. It is often misused to make predictions about an event being "due" because it has happened less frequently than expected in a small sample size.


## 3.2.3 How Large is Large in CLT?

**Key points**

* The sample size required for the Central Limit Theorem and Law of Large Numbers to apply differs based on the probability of success.
  + If the probability of success is high, then relatively few observations are needed.
  + As the probability of success decreases, more observations are needed.
* If the probability of success is extremely low, such as winning a lottery, then the Central Limit Theorem may not apply even with extremely large sample sizes. The normal distribution is not a good approximation in these cases, and other distributions such as the Poisson distribution (not discussed in these courses) may be more appropriate.


## 3.2.4 Assessment

### Exercise 1. American Roulette probability of winning money
The exercises in the previous chapter explored winnings in American roulette. In this chapter of exercises, we will continue with the roulette example and add in the Central Limit Theorem.
In the previous chapter of exercises, you created a random variable S that is the sum of your winnings after betting on green a number of times in American Roulette.
What is the probability that you end up winning money if you bet on green 100 times?

*Instructions*

* Execute the sample code to determine the expected value avg and standard error se as you have done in previous exercises.
* Use the pnorm function to determine the probability of winning money.

*Code*
```{r}
library(gtools)
library(tidyverse)
options(digits = 3)    # report 3 significant digits

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- 2 / 38

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Define the number of bets using the variable 'n'
n <- 100

# Calculate 'avg', the expected outcome of 100 spins if you win $17 when the ball lands on green and you lose $1 when the ball doesn't land on green
avg <- n * (17*p_green + -1*p_not_green)
avg

# Compute 'se', the standard error of the sum of 100 outcomes
se <- sqrt(n) * (17 - -1)*sqrt(p_green*p_not_green)
se

# Using the expected value 'avg' and standard error 'se', compute the probability that you win money betting on green 100 times.
1-pnorm(0, avg, se)
```

### Exercise 2. American Roulette Monte Carlo simulation
Create a Monte Carlo simulation that generates 10,000 outcomes of S, the sum of 100 bets.
Compute the average and standard deviation of the resulting list and compare them to the expected value (-5.263158) and standard error (40.19344) for S that you calculated previously.

*Instructions*

* Use the replicate function to replicate the sample code for B <- 10000 simulations.
* Within replicate, use the sample function to simulate n <- 100 outcomes of either a win (17) or a loss (-1) for the bet. Use the order c(17, -1) and corresponding probabilities. Then, use the sum function to add up the winnings over all iterations of the model. Make sure to include sum or DataCamp may crash with a "Session Expired" error.
* Use the mean function to compute the average winnings.
* Use the sd function to compute the standard deviation of the winnings.

*Code*
```{r}
# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- 2 / 38

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1-p_green

# Define the number of bets using the variable 'n'
n <- 100

# The variable `B` specifies the number of times we want the simulation to run. Let's run the Monte Carlo simulation 10,000 times.
B <- 10000

# Use the `set.seed` function to make sure your answer matches the expected result after random sampling.
set.seed(1)

# Create an object called `S` that replicates the sample code for `B` iterations and sums the outcomes.
S <- replicate(B, {
  x <- sample(c(17,-1,-1), n, replace=TRUE, prob=c(p_green, p_not_green/2, p_not_green/2))
  sum(x)
})

# Compute the average value for 'S'
mean(S)

# Calculate the standard deviation of 'S'
sd(S)
```


### Exercise 3. American Roulette Monte Carlo vs CLT
In this chapter, you calculated the probability of winning money in American roulette using the CLT.
Now, calculate the probability of winning money from the Monte Carlo simulation. The Monte Carlo simulation from the previous exercise has already been pre-run for you, resulting in the variable S that contains a list of 10,000 simulated outcomes.

*Instructions*

*Code*
```{r}
# Calculate the proportion of outcomes in the vector `S` that exceed $0
mean(S>0)
```

### Exercise 4. American Roulette Monte Carlo vs CLT comparison
The Monte Carlo result and the CLT approximation for the probability of losing money after 100 bets are close, but not that close. What could account for this?

### Exercise 5. American Roulette average winnings per bet
Now create a random variable Y that contains your average winnings per bet after betting on green 10,000 times.

*Instructions*

* Run a single Monte Carlo simulation of 10,000 bets using the following steps. (You do not need to replicate the sample code.)
* Specify n as the number of times you want to sample from the possible outcomes.
* Use the sample function to return n values from a vector of possible values: winning $17 or losing $1. Be sure to assign a probability to each outcome and indicate that you are sampling with replacement.
* Calculate the average result per bet placed using the mean function.

*Code*
```{r}
# Use the `set.seed` function to make sure your answer matches the expected result after random sampling.
set.seed(1)

# Define the number of bets using the variable 'n'
n <- 10000

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- 2 / 38

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1 - p_green

# Create a vector called `X` that contains the outcomes of `n` bets
X <- sample(c(17,-1), n, replace=TRUE, prob=c(p_green, p_not_green))

# Define a variable `Y` that contains the mean outcome per bet. Print this mean to the console.
Y <- mean(X)
Y
```

### Exercise 6. American Roulette per bet expected value
What is the expected value of Y, the average outcome per bet after betting on green 10,000 times?

*Instructions*

*Code*
```{r}
# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- 2 / 38

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1 - p_green

# Calculate the expected outcome of `Y`, the mean outcome per bet in 10,000 bets
n <- 10000
p_green * 17 + p_not_green * (-1)
```

### Exercise 7. American Roulette per bet standard error
What is the standard error of Y, the average result of 10,000 spins?

*Instructions*

*Code*
```{r}
# Define the number of bets using the variable 'n'
n <- 10000

# Assign a variable `p_green` as the probability of the ball landing in a green pocket
p_green <- 2 / 38

# Assign a variable `p_not_green` as the probability of the ball not landing in a green pocket
p_not_green <- 1 - p_green

# Compute the standard error of 'Y', the mean outcome per bet from 10,000 bets.
n <- 10000
1/sqrt(n) * abs(17 - (-1)) * sqrt(p_green * p_not_green)
```

### Exercise 8. American Roulette winnings per game are positive
What is the probability that your winnings are positive after betting on green 10,000 times?

*Instructions*

* Execute the code that we wrote in previous exercises to determine the average and standard error.
* Use the pnorm function to determine the probability of winning more than $0.

*Code*
```{r}
1-pnorm(0, avg, se)
```

### Exercise 9. American Roulette Monte Carlo again
Create a Monte Carlo simulation that generates 10,000 outcomes of S, the average outcome from 10,000 bets on green.
Compute the average and standard deviation of the resulting list to confirm the results from previous exercises using the Central Limit Theorem.

*Instructions*

* Use the replicate function to model 10,000 iterations of a series of 10,000 bets.
* Each iteration inside replicate should simulate 10,000 bets and determine the average outcome of those 10,000 bets. If you forget to take the mean, DataCamp will crash with a "Session Expired" error.
* Find the average of the 10,000 average outcomes. Print this value to the console.
* Compute the standard deviation of the 10,000 simulations. Print this value to the console.

*Code*
```{r}
## Make sure you fully follow instructions, including printing values to the console and correctly running the `replicate` loop. If not, you may encounter "Session Expired" errors.

# The variable `n` specifies the number of independent bets on green
n <- 10000

# The variable `B` specifies the number of times we want the simulation to run
B <- 10000

# Use the `set.seed` function to make sure your answer matches the expected result after random number generation
set.seed(1)

# Generate a vector `S` that contains the the average outcomes of 10,000 bets modeled 10,000 times
S <- replicate(B, {
  x <- sample(c(17,-1), n, replace=TRUE, prob=c(p_green, p_not_green))
  mean(x)
})

# Compute the average of `S`
mean(S)

# Compute the standard deviation of `S`
sd(S)

```

### Exercise 10. American Roulette comparison
In a previous exercise, you found the probability of winning more than $0 after betting on green 10,000 times using the Central Limit Theorem. Then, you used a Monte Carlo simulation to model the average result of betting on green 10,000 times over 10,000 simulated series of bets.

What is the probability of winning more than $0 as estimated by your Monte Carlo simulation? The code to generate the vector S that contains the the average outcomes of 10,000 bets modeled 10,000 times has already been run for you.

*Instructions*

* Calculate the probability of winning more than $0 in the Monte Carlo simulation from the previous exercise.

* You do not need to run another simulation: the results of the simulation are in your workspace as the vector S

*Code*
```{r}
# Compute the proportion of outcomes in the vector 'S' where you won more than $0
mean(S>0)
```
