---
title: "Section 3: Random Variables, Sampling Models, and the Central Limit Theorem"
output: html_notebook
---

# 3.0 Overview


Section 3 introduces you to Random Variables, Sampling Models, and the Central Limit Theorem.

Section 3 is divided into two parts:

* Random Variables and Sampling Models
* The Central Limit Theorem.

After completing Section 3, you will:

* understand what random variables are, how to generate them, and the correct mathematical notation to use with them.
* be able to use sampling models to estimate characteristics of a larger population.
* be able to explain the difference between a distribution and a probability distribution.
* understand the Central Limit Theorem and the law of large numbers.

There are 2 assignments that use the DataCamp platform for you to practice your coding skills as well as a set of questions on the edX platform at the end of Section 3.  

This section corresponds to the following section of the course textbook.  

We encourage you to use R to interactively test out your answers and further your learning.

```{r}
library(gtools)
library(tidyverse)
options(digits = 3)    # report 3 significant digits
```

# 3.3 Assessment

## 3.3.1 Questions 1 and 2: SAT testing

The SAT is a standardized college admissions test used in the United States. The following two multi-part questions will ask you some questions about SAT testing.

An old version of the SAT college entrance exam had a -0.25 point penalty for every incorrect answer and awarded 1 point for a correct answer. The quantitative test consisted of 44 multiple-choice questions each with 5 answer choices. Suppose a student chooses answers by guessing for all questions on the test.

```{r}
p <- 1/5
a <- 1
b <- -1/4
n <- 44

mu <- a*p + b*(1-p)
 
# 1.a:
# What is the probability of guessing correctly for one question?
p

# 1.e:
# Use the Central Limit Theorem to determine the probability that a guessing student scores 8 points or higher on the test.
1-pnorm(8, n*mu, sqrt(n)*abs(a-b)*sqrt(p*(1-p)))

# Set the seed to 21, then run a Monte Carlo simulation of 10,000 students guessing on the test.
set.seed(1)
B <- 10000

S <- replicate(B, {
  points <- sample(c(a,b), n, replace=TRUE, prob=c(p,1-p))
  sum(points)
})

# 1.f:
# What is the probability that a guessing student scores 8 points or higher?
mean(S>=8)

```

The SAT was recently changed to reduce the number of multiple choice options from 5 to 4 and also to eliminate the penalty for guessing.

```{r}
p <- 1/4
a <- 1
b <- 0
n <- 44

mu <- a*p + b*(1-p)

# 2.a:
# Suppose that the number of multiple choice options is 4 and that there is no penalty for guessing - that is, an incorrect question gives a score of 0.
# What is the expected value of the score when guessing on this new test?
mu <- a*p + b*(1-p)
n* mu

# 2.b:
# What is the probability of scoring over 30 when guessing?
1-pnorm(30, n*mu, sqrt(n)*abs(a-b)*sqrt(p*(1-p)))

# 2.c:
# Consider a range of correct answer probabilities p <- seq(0.25, 0.95, 0.05) representing a range of student skills.
# What is the lowest p such that the probability of scoring over 35 exceeds 80%?
test <- function(p) {
  mu <- a*p + b*(1-p)
  1-pnorm(35, n*mu, sqrt(n)*abs(a-b)*sqrt(p*(1-p)))
}

p_ <- seq(0.25, 0.95, 0.05)
res <- sapply(p_, test)

min(p_[which(res > 0.8)])
```

## 3.3.2 Betting on Roulette

A casino offers a House Special bet on roulette, which is a bet on five pockets (00, 0, 1, 2, 3) out of 38 total pockets. The bet pays out 6 to 1. In other words, a losing bet yields -\$1 and a successful bet yields \$6. A gambler wants to know the chance of losing money if he places 500 bets on the roulette House Special.

```{r}
p <- 5/38
a <- 6
b <- -1
n <- 500
```

```{r}

# 3.a:
# expected value of the payout for one bet?
mu <- p*a + (1-p) * b
mu

# 3.b:
# stddev of the payout for one bet?
sigma <- abs(a-b) * sqrt(p * (1-p))
sigma

# 3.c:
# expected value of the average payout over 500 bets?
n * mu

# 3.d:
# standard error of the average payout over 500 bets?
sigma/sqrt(n)

# 3.e:
# expected value of the sum of 500 bets?
n * mu

# 3.f:
# standard error of the sum of 500 bets?
sqrt(n) * sigma

# 3.g:
# calculate the probability of losing money over 500 bets
pnorm(0, n*mu, sqrt(n) * sigma)
```