---
title: 'Data Science: Wrangling - HarvardX: PH125.6x'
author: 'Luiz Cunha'
date: '2019-08-12'
output: html_notebook
---

# Section 1: Data Import

## Overview

In the **Data Import** section, you will learn how import data into R.  
After completing this section, you will be able to:

* **Import** data from spreadsheets.
* Identify and set your **working directory** and specify the path to a file.
* Use the *readr* and *readxl* packages to import spreadsheets.
* Use **R-base functions** to import spreadsheets.
* **Download** files from the internet using R.

## 1.1 Data Import

### 1.1.1 Importing Spreadsheets

**Key points**

* Many datasets are stored in spreadsheets. A spreadsheet is essentially a file version of a data frame with rows and columns.
* Spreadsheets have rows separated by returns and columns separated by a delimiter. The most common delimiters are comma, semicolon, white space and tab.
* Many spreadsheets are raw text files and can be read with any basic text editor. However, some formats are proprietary and cannot be read with a text editor, such as Microsoft Excel files (.xls).
* Most import functions assume that the first row of a spreadsheet file is a header with column names. To know if the file has a header, it helps to look at the file with a text editor before trying to import it.


### 1.1.2 Paths and the Working Directory

**Key points**

* The working directory is where R looks for files and saves files by default.
* See your working directory with getwd. Change your working directory with setwd.
* We suggest you create a directory for each project and keep your raw data inside that directory.
* Use the file.path function to generate a full path from a relative path and a file name. * Use file.path instead of paste because file.path is aware of your operating system and will use the correct slashes to navigate your machine.
* The file.copy function copies a file to a new path.

**Code**
```{r eval = FALSE}

# see working directory
wd <- getwd()

# change your working directory
setwd(wd)

# set path to the location for raw data files in the dslabs package and list files
path <- system.file("extdata", package="dslabs")
list.files(path)

# generate a full path to a file
filename <- "murders.csv"
fullpath <- file.path(path, filename)
fullpath

# copy file from dslabs package to your working directory
file.copy(fullpath, getwd())
# check if the file exists
file.exists(filename)
```


### 1.1.3 The readr and readxl Packages

**Key points**

* **readr** is the **tidyverse** library that includes functions for reading data stored in text file spreadsheets into R. Functions in the package include read_csv, read_tsv, read_delim and more. These differ by the delimiter they use to split columns:
  + **read_table**: (usually .txt) white space separated values
  + **read_csv**: (usually .csv) comma separated values
  + **read_csv2**: (usually .csv) semicolonscomma separated values
  + **read_tsv**: (usually .tsv) tab separated values
  + **read_delim**: (usually .txt) general .txt file format, must define delimiter
* The **readxl** package provides functions to read Microsoft Excel formatted files.
* The *excel_sheets* function gives the names of the sheets in the Excel file. These names are passed to the sheet argument for the **readxl** functions *read_excel, read_xls and read_xlsx*.
  + **read_excel**: (usually .xls .xlsx) auto detect the format
  + **read_xls**: (usually .xls) original format
  + **read_xlx**: (usually .xlsx) new format
* The *read_lines* function shows the first few lines of a file in R.

**Code**
```{r}
library(dslabs)
library(tidyverse)    # includes readr
library(readxl)

# inspect the first 3 lines
filename <- "murders.csv" 
read_lines(filename, n_max = 3)

# read file in CSV format
dat <- read_csv(filename)

#read using full path
fullpath <- file.path(getwd(), filename)
dat <- read_csv(fullpath)
head(dat)

#Ex:
path <- system.file("extdata", package = "dslabs")
files <- list.files(path)
files

filename <- "murders.csv"
filename1 <- "life-expectancy-and-fertility-two-countries-example.csv"
filename2 <- "fertility-two-countries-example.csv"
dat=read_csv(file.path(path, filename))
dat1=read_csv(file.path(path, filename1))
dat2=read_csv(file.path(path, filename2))
```


### 1.1.4 Importing Data Using R-base Functions

**Key point**

R-base import functions *(read.csv, read.table, read.delim)* generate *data frames* rather than tibbles and character variables are converted to factors. This can be avoided by setting the argument *stringsAsFactors=FALSE*.

**Code**
```{r}
# filename is defined in the previous video
# read.csv converts strings to factors
dat2 <- read.csv(filename)
class(dat2)
class(dat2$abb)
class(dat2$region)
```


### 1.1.5 Downloading Files from the Internet

**Key points**

* The **read_csv** function and other import functions can read a URL directly.
* If you want to have a local copy of the file, you can use *download.file*.
* *tempdir* creates a directory with a name that is very unlikely not to be unique.
* *tempfile* creates a character string that is likely to be a unique filename.

**Code**
```{r}
url <- "https://raw.githubusercontent.com/rafalab/dslabs/master/inst/extdata/murders.csv"
dat <- read_csv(url)
download.file(url, "murders.csv")
tempfile()
tmp_filename <- tempfile()
download.file(url, tmp_filename)
dat <- read_csv(tmp_filename)
file.remove(tmp_filename)
```


### 1.1.6 Assessment Part 1: Data Import

#### Question 15
Check the documentation for the readr function you chose in the previous question to learn about its arguments. Determine which arguments you need to the file from the previous question:

```{r}
url <- "http://mlr.cs.umass.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data"
```

Does this file have a header row? Does the readr function you chose need any additional arguments to import the data correctly?

```{r}
# A: No, there is no header. The col_names=FALSE argument is necessary.
dat <- read_csv(url, col_names = FALSE)

# Inspect the data
head(dat,5)

# How many rows, cols are in the dataset?
dim(dat)
```

#### Question 17
Check the documentation for the readr function you chose in the previous question to learn about its arguments. Determine which arguments you need to the file from the previous question:

```{r}
url <- "ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_annmean_mlo.txt"
```

Does this file have a header row? Does the readr function you chose need any additional arguments to import the data correctly?

```{r}
# Download the file into a tempfile in working directory
tmp_filename <- tempfile(tmpdir=getwd())
download.file(url, tmp_filename)

# read the first lines of the file
read_lines(tmp_filename, n_max = 60)

# file fields separated by white space => use read_table
# top 56 lines to skip
# col '#' to be removed
co2_mauna_loa <- read_table(tmp_filename, col_names = TRUE, skip=56)
colnames(co2_mauna_loa)
co2_mauna_loa <- select(co2_mauna_loa, c(year, mean, unc))

# delete the temp file (once data loaded)
file.remove(tmp_filename)

# check the data format is clean
head(co2_mauna_loa)

# check the tibble dimensions
dim(co2_mauna_loa)
```


