---
title: "Section3: Summarizing with dplyr"
output: html_notebook
---

# Summarizing with dplyr

Section 3 introduces you to summarizing with dplyr.

After completing Section 3, you will:  
* understand the importance of summarizing data in exploratory data analysis.  
* be able to use the "summarize" verb in dplyr to facilitate summarizing data.  
* be able to use the "group_by" verb in dplyr to facilitate summarizing data.  
* be able to access values using the dot placeholder.  
* be able to use "arrange" to examine data after sorting.  


## dplyr

**Key points**

* summarize from the dplyr/tidyverse package computes summary statistics from the data frame. It returns a data frame whose column names are defined within the function call.  
* summarize can compute any summary function that operates on vectors and returns a single value, but it cannot operate on functions that return multiple values.  
* Like most dplyr functions, summarize is aware of variable names within data frames and can use them directly.

**Code**

```{r}
library(tidyverse)
library(dslabs)
data(heights)

# compute average and standard deviation for males
s <- heights %>%
    filter(sex == "Male") %>%
    summarize(average = mean(height), standard_deviation = sd(height))

# access average and standard deviation from summary table
s$average
s$standard_deviation

# compute median, min and max
heights %>%
    filter(sex == "Male") %>%
    summarize(median = median(height),
                       minimum = min(height),
                       maximum = max(height))

# alternative way to get min, median, max in base R
quantile(heights$height, c(0, 0.5, 1))

# NB !!! generates an error: summarize can only take functions that return a single value
#heights %>%
#    filter(sex == "Male") %>%
#    summarize(range = quantile(height, c(0, 0.5, 1)))
```


## The Dot Placeholder

Note that a common replacement for the dot operator is the pull function

**Key points**

* The dot operator allows you to access values stored in data that is being piped in using the %>% character. The dot is a placeholder for the data being passed in through the pipe.  
* The dot operator allows dplyr functions to return single vectors or numbers instead of only data frames.  
* us_murder_rate %>% .$rate is equivalent to us_murder_rate$rate.  
* Note that an equivalent way to extract a single column using the pipe is us_murder_rate %>% pull(rate). The pull function will be used in later course material.

**Code**

```{r}
library(tidyverse)
library(dslabs)
data(murders)

murders <- murders %>% mutate(murder_rate = total/population*100000)
summarize(murders, mean(murder_rate))

# calculate US murder rate, generating a data frame
us_murder_rate <- murders %>%
    summarize(rate = sum(total) / sum(population) * 100000)
us_murder_rate

# extract the numeric US murder rate with the dot operator
us_murder_rate %>% .$rate

# calculate and extract the murder rate with one pipe
us_murder_rate <- murders %>%
      summarize(rate = sum(total) / sum(population * 100000)) %>%
      .$rate
```


## Group By

**Key points**

* The group_by function from *dplyr*  converts a data frame to a grouped data frame, creating groups using one or more variables.  
* *summarize* and some other *dplyr* functions will behave differently on grouped data frames.  
* Using *summarize* on a grouped data frame computes the summary statistics for each of the separate groups.

**Code**

```{r}

# libraries and data
library(tidyverse)
library(dslabs)
data(heights)
data(murders)

# compute separate average and standard deviation for male/female heights
heights %>%
    group_by(sex) %>%
    summarize(average = mean(height), standard_deviation = sd(height))

# compute median murder rate in 4 regions of country
murders <- murders %>%
      mutate(murder_rate = total/population * 100000)

murders %>%
      group_by(region) %>%
      summarize(median_rate = median(murder_rate))
```


## Sorting Data Tables

**Key points**

* The arrange function from dplyr sorts a data frame by a given column.  
* By default, arrange sorts in ascending order (lowest to highest). To instead sort in descending order, use the function desc inside of arrange.  
* You can arrange by multiple levels: within equivalent values of the first level, observations are sorted by the second level, and so on.  
* The top_n function shows the top results ranked by a given variable, but the results are not ordered.  
* You can combine top_n with arrange to return the top results in order.

**Code**

```{r}
# libraries and data
library(tidyverse)
library(dslabs)
data(murders)

# set up murders object
murders <- murders %>%
    mutate(murder_rate = total/population * 100000)

# arrange by population column, smallest to largest
murders %>% arrange(population) %>% head()

# arrange by murder rate, smallest to largest
murders %>% arrange(murder_rate) %>% head()

# arrange by murder rate in descending order
murders %>% arrange(desc(murder_rate)) %>% head()

# arrange by region alphabetically, then by murder rate within each region
murders %>% arrange(region, murder_rate) %>% head()

# show the top 10 states with highest murder rate, not ordered by rate
murders %>% top_n(10, murder_rate)

# show the top 10 states with highest murder rate, ordered by rate
murders %>% arrange(desc(murder_rate)) %>% top_n(10)
```
