---
title: 'Data Science: MachineR Basics - HarvardX: PH125.1x'
author: 'Luiz Cunha'
date: '2019-08-19'
output: html_notebook
---

```{r include=FALSE}
# Set knitr options for knitting code into the report:
# - echo=TRUE: print out code (echo), though any results/output would still be displayed
# - cache=TRUE: Save results so that code blocks aren't re-run unless code changes (cache),
# _or_ a relevant earlier code block changed (autodep), but don't re-run if the
# only thing that changed was the comments (cache.comments)
# - message, warning = FALSE: Don't clutter R output with messages or warnings
# This _will_ leave error messages showing up in the knitted report
# - results="hide": hide the results/output (but here the code would still be displayed)
# - include=FALSE:  chunk is evaluated, but neither the code nor its output is displayed
# - eval=FALSE: chunk is not evaluated
knitr::opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE)
```

# Section 4: Programming Basics

## Overview

Section 4 introduces you to general programming features like 'if-else' and 'for loop' commands so that you can write your own functions to perform various operations on datasets.

In Section 4.1, you will:

* Understand some of the programming capabilities of R. 

In Section 4.2, you will:

* Use basic conditional expressions to perform different operations. 
* Check if any or all elements of a logical vector are TRUE. 

In Section 4.3, you will:

* Define and call functions to perform various operations.
* Pass arguments to functions, and return variables/objects from functions.

In Section 4.4, you will:

* Use for-loops to perform repeated operations. 
* Articulate in-built functions of R that you could try for yourself. 

## 4.1 Introduction to Programming in R
...

## 4.2 Conditionals

### 4.2.1 Basics Conditionals

**Key Points**

* The most common conditional expression in programming is an if-else statement, which has the form "if [condition], perform [expression], else perform [alternative expression]".
* The **ifelse()** function works similarly to an if-else statement, but it is particularly useful since it **works on vectors by examining each element** of the vector and returning a corresponding answer accordingly.
* The **any()** function takes a vector of logicals and returns true if any of the entries are true.
* The **all()** function takes a vector of logicals and returns true if all of the entries are true.

**Code**
```{r}
# an example showing the general structure of an if-else statement
a <- 0
if(a!=0){
  print(1/a)
} else{
  print("No reciprocal for 0.")
}

# an example that tells us which states, if any, have a murder rate less than 0.5
library(dslabs)
data(murders)
murder_rate <- murders$total / murders$population*100000
ind <- which.min(murder_rate)
if(murder_rate[ind] < 0.5){
  print(murders$state[ind]) 
} else{
  print("No state has murder rate that low")
}

# changing the condition to < 0.25 changes the result
if(murder_rate[ind] < 0.25){
  print(murders$state[ind]) 
} else{
  print("No state has a murder rate that low.")
}

# the ifelse() function works similarly to an if-else conditional
a <- 0
ifelse(a > 0, 1/a, NA)

# the ifelse() function is particularly useful on vectors
a <- c(0,1,2,-4,5)
result <- ifelse(a > 0, 1/a, NA)

# the ifelse() function is also helpful for replacing missing values
data(na_example)
no_nas <- ifelse(is.na(na_example), 0, na_example) 
sum(is.na(no_nas))

# the any() and all() functions evaluate logical vectors
z <- c(TRUE, TRUE, FALSE)
any(z)
all(z)
```


## 4.3 Functions

### 4.3.1 Basic Functions

**Key points**

* The R function, called **function()** tells R you are about to define a new function.
* **Functions are objects**, so **must be assigned a variable name** with the arrow operator.
* The general way to define functions is: 
  + 1. decide the function name, which will be an object, 
  + 2. type function() with your function's arguments in parentheses, 
  + 3. write all the operations inside brackets.
* Variables defined inside a function are not saved in the workspace.

**Code**
```{r}
# example of defining a function to compute the average of a vector x
avg <- function(x){
  s <- sum(x)
  n <- length(x)
  s/n
}

# we see that the above function and the pre-built R mean() function are identical
x <- 1:100
identical(mean(x), avg(x))

# variables inside a function are not defined in the workspace
s <- 3
avg(1:10)
s

# the general form of a function
my_function <- function(VARIABLE_NAME){
#  perform operations on VARIABLE_NAME and calculate VALUE
#  VALUE
}

# functions can have multiple arguments as well as default values
avg <- function(x, arithmetic = TRUE){
  n <- length(x)
  ifelse(arithmetic, sum(x)/n, prod(x)^(1/n))
}
```


## 4.4 For Loops

### 4.4.1 For Loops

**Key points**

* **For-loops** perform the same task over and over while changing the variable.  They let us define the range that our variable takes, and then changes the value with each loop and evaluates the expression every time inside the loop.
* The general form of a for-loop is: "For i in [some range], do operations".  This i changes across the range of values and the operations assume i is a value you're interested in computing on.
* At the end of the loop, the value of i is the last value of the range.

**Code**
```{r}
# creating a function that computes the sum of integers 1 through n
compute_s_n <- function(n){
  x <- 1:n
  sum(x)
}

# a very simple for-loop
for(i in 1:5){
  print(i)
}

# a for-loop for our summation
m <- 25
s_n <- vector(length = m) # create an empty vector
for(n in 1:m){
  s_n[n] <- compute_s_n(n)
}

# creating a plot for our summation function
n <- 1:m
plot(n, s_n)

# a table of values comparing our function to the summation formula
head(data.frame(s_n = s_n, formula = n*(n+1)/2))

# overlaying our function with the summation formula
plot(n, s_n)
lines(n, n*(n+1)/2)
```

### 4.4.2 Other Functions
...

**Key points**

* For-loops rarely used, instead use: 
  + apply, sapply, tapply, mapply
  + split(x ,f): split divides the data in the vector x into the groups defined by f
  + cut(x, breaks): cut divides the range of x into intervals and codes the values in x according to which interval they fall
  + quantile
  + reduce
  + identical: the safe and reliable way to test two objects for being exactly equal
  + unique
  
**Code**
```{r}
# Split example
split(1:10, 1:2)

# Cut example
aaa <- c(1,2,3,4,5,2,3,4,5,6,7)
cut(aaa, 3)
```

