---
title: "Section 2: Continuous Probability"
output: html_notebook
---

# 2.2 Assessment


## 2.2.1 Questions 1 and 2: ACT scores, part 1

The ACT is a standardized college admissions test used in the United States. The four multi-part questions in this assessment all involve simulating some ACT test scores and answering probability questions about them.

For the three year period 2016-2018, ACT standardized test scores were approximately normally distributed with a mean of 20.9 and standard deviation of 5.7. (Real ACT scores are integers between 1 and 36, but we will ignore this detail and use continuous values instead.)

First we'll simulate an ACT test score dataset and answer some questions about it.

Set the seed to 16, then use rnorm to generate a normal distribution of 10000 tests with a mean of 20.9 and standard deviation of 5.7. Save these values as act_scores. You'll be using this dataset throughout these four multi-part questions.

**Code**

```{r}
library(gtools)
library(tidyverse)
options(digits = 3)    # report 3 significant digits

set.seed(16)
n <- 10000
mu <- 20.9
sigma <- 5.7

act_scores <- rnorm(n, mu, sigma)

# 1.a: mean?
mean(act_scores)

# 1.b: stddev?
sd(act_scores)

# 1.c:
# A perfect score is 36 or greater (the maximum reported score is 36).
# how many perfect scores are there out of 10,000 simulated tests?
sum(act_scores >= 36)

# 1.d:
# what is the probability of an ACT score greater than 30?
sum(act_scores >= 30) / n

# 1.e:
# what is the probability of an ACT score less than or equal to 10?
sum(act_scores <= 10) / n
  
```


## 2.2.2 Questions 3 and 4: ACT scores, part 2

In this 3-part question, you will convert raw ACT scores to Z-scores and answer some questions about them.

Convert act_scores to Z-scores. Recall from Data Visualization (the second course in this series) that to standardize values (convert values into Z-scores, that is, values distributed with a mean of 0 and standard deviation of 1), you must subtract the mean and then divide by the standard deviation. Use the mean and standard deviation of act_scores, not the original values used to generate random test scores.

```{r}
z_act <- (act_scores - mean(act_scores)) / sd(act_scores)

# 3.a: z > 2
mean(z_act >=2)

# 3.b: score at z=2
mean(act_scores) + 2* sd(act_scores)

# 3.c:
# A Z-score of 2 corresponds roughly to the 97.5th percentile.
# Use qnorm to determine the 97.5th percentile of normally distributed data with the mean and standard deviation observed in act_scores.
# What is the 97.5th percentile?
qnorm(0.975, mean(act_scores), sd(act_scores))

# 4:
# Write a function that takes a value and produces the probability of an ACT score less than or equal to that value (the CDF). Apply this function to the range 1 to 36.
f <- function(x) {
  mean(act_scores<=x)
}

# 4.a:
# What is the minimum score such that the probability of that score or lower is at least .95?
res <- sapply(1:36,f)
data.frame(ind = 1:36, res = res) %>%
  filter(res > 0.95) %>%
  .$ind

# 4.b:
# Use qnorm to determine the expected 95th percentile, the value for which the probability of receiving that score or lower is 0.95, given a mean score of 20.9 and standard deviation of 5.7.
qnorm(0.95, mu, sigma)

# 4.c:
# Make a vector containing the quantiles for p <- seq(0.01, 0.99, 0.01), the 1st through 99th percentiles of the act_scores data. Save these as sample_quantiles.
# In what percentile is a score of 26?
p <- seq(0.01, 0.99, 0.01)
sample_quantiles <- quantile(act_scores, p)
names(sample_quantiles[max(which(sample_quantiles < 26))])

# 4.d:
# Make a corresponding set of theoretical quantiles using qnorm over the interval p <- seq(0.01, 0.99, 0.01) with mean 20.9 and standard deviation 5.7. Save these as theoretical_quantiles. Make a QQ-plot
theoretical_quantiles <- qnorm(p, mu, sigma)
qplot(theoretical_quantiles, sample_quantiles) +
  geom_abline()
```
