---
title: "Data Science: Inference and Modeling - HarvardX: PH125.4x"
output: html_notebook
---

# Section 1: Parameters and Estimates

## 1.1 Overview

Section 1 introduces you to parameters and estimates.

After completing Section 1, you will be able to:

* Understand how to use a sampling model to perform a poll.
* Explain the terms population, parameter, and sample as they relate to statistical inference.
* Use a sample to estimate the population proportion from the sample average.
* Calculate the expected value and standard error of the sample average. 


## 1.2 Parameters and Estimates

### 1.2.1 Sampling Model Parameters and Estimates

**Key points**

* The task of statistical inference is to estimate an unknown population parameter using observed data from a sample.
* In a sampling model, the collection of elements in the urn is called the population.
* A *parameter* is a number that summarizes data for an entire *population*.
* A *sample* is observed data from a subset of the population.
* An *estimate* is a summary of the observed data about a parameter that we believe is informative. It is a data-driven guess of the population parameter.
* We want to predict the proportion of the blue beads in the urn, the parameter *p*. The proportion of red beads in the urn is *1-p* and the spread is *2p-1*.
* The sample proportion is a random variable. Sampling gives random results drawn from the population distribution.

**Code: Function for taking a random draw from a specific urn**

The dslabs package includes a function for taking a random draw of size  ????  from the urn described in the video:
```{r}
library(tidyverse)
library(dslabs)
take_poll(25)    # draw 25 beads
```

### 1.2.2 The Sample Average

**Key points**

* Many common data science tasks can be framed as estimating a parameter from a sample.
* We illustrate statistical inference by walking through the process to estimate *p*. From the estimate of *p*, we can easily calculate an estimate of the spread, *2p-1*.
* Consider the random variable *X* that is 1 if a blue bead is chosen and 0 if a red bead is chosen. The proportion of blue beads in *N* draws is the average of the draws $X_1,...,X_n$.
* $\bar{X}$ is the *sample average*. In statistics, a bar on top of a symbol denotes the average. $\bar{X}$ is a random variable because it is the average of random draws - each time we take a sample, $\bar{X}$ is different.
$$ \bar{X} = \frac{X_1+X_2...+X_n}{N}$$
* The number of blue beads drawn in N draws, $N\bar{X}$, is *N* times the proportion of values in the urn. However, we do not know the true proportion: we are trying to estimate this parameter *p*.


### 1.2.3 Polling vs Forecasting

**Key points**

* A poll taken in advance of an election estimates *p* for that moment, not for election day.
* In order to predict election results, forecasters try to use early estimates of *p* to predict *p* on election day. We discuss some approaches in later sections.


### 1.2.4 Properties of our Estimate

***Key points**

* When interpreting values of $\bar{X}$, it is important to remember that $\bar{X}$ is a random variable with an expected value and standard error that represents the sample proportion of positive events.
* The expected value of $\bar{X}$ is the parameter of interest p. This follows from the fact that $\bar{X}$ is the sum of independent draws of a random variable times a constant *1/N* .
$$ E[\bar{X}] = p $$
* As the number of draws $N$ increases, the standard error of our estimate $\bar{X}$ decreases. The standard error of the average of $\bar{X}$ over $N$ draws is:
$$ SE[\bar{X}] = \sqrt{\frac{p(1-p)}{N}} $$
* In theory, we can get more accurate estimates of $p$ by increasing $N$ . In practice, there are limits on the size of $N$ due to costs, as well as other factors we discuss later.
* We can also use other random variable equations to determine the expected value of the sum of draws $E[\bar{X}]$ and standard error of the sum of draws $SE[\bar{X}]$.
$$ E[S] = Np $$
$$ SE[S] = \sqrt{Np(1-p)} $$


### 1.2.5 Assessment

**Exercise 5. se versus p**
Write a line of code that calculates the standard error se of a sample average when you poll 25 people in the population. Generate a sequence of 100 proportions of Democrats p that vary from 0 (no Democrats) to 1 (all Democrats).  
Plot se versus p for the 100 different proportions.

```{r}
# `N` represents the number of people polled
N <- 25

# Create a variable `p` that contains 100 proportions ranging from 0 to 1 using the `seq` function
p <- seq(0,1,length.out=100)

# Create a variable `se` that contains the standard error of each sample average
se <- sqrt(p*(1-p)/N)

# Plot `p` on the x-axis and `se` on the y-axis
plot(p, se)
```

**Exercise 6. Multiple plots of se versus p**
Using the same code as in the previous exercise, create a for-loop that generates three plots of p versus se when the sample sizes equal N=25, N=100, and N=1000.

```{r}
# The vector `p` contains 100 proportions of Democrats ranging from 0 to 1 using the `seq` function
p <- seq(0, 1, length = 100)

# The vector `sample_sizes` contains the three sample sizes
sample_sizes <- c(25, 100, 1000)

# Write a for-loop that calculates the standard error `se` for every value of `p` for each of the three samples sizes `N` in the vector `sample_sizes`. Plot the three graphs, using the `ylim` argument to standardize the y-axis across all three plots.
for (N in sample_sizes) {
   # Create a variable `se` that contains the standard error of each sample average
    se <- sqrt(p*(1-p)/N)
    # Plot `p` on the x-axis and `se` on the y-axis
    plot(p, se, ylim=c(0, 0.1))
}
```