---
title: "Data Science: Visualization"
author: Luiz Cunha
output: html_notebook
---

# Section 2. Introduction to ggplot2

## Overview

In Section 2, you will learn how to create data visualizations in R using ggplot2.

After completing Section 2, you will:

* be able to use ggplot2 to create data visualizations in R.
* be able to explain what the data component of a graph is.
* be able to identify the geometry component of a graph and know when to use which type of geometry.
* be able to explain what the aesthetic mapping component of a graph is.
* be able to understand the scale component of a graph and select an appropriate scale component to use.

```{r}
library(tidyverse)
```

## 2.1 Basics of ggplot2

## 2.1.1 ggplot2

**Key points and notes**

* Throughout the series, we will create plots with the **ggplot2** package. ggplot2 is part of the tidyverse, which you can load with library(tidyverse).
* Note that you can also load ggplot2 alone using the command *library(ggplot2)*, instead of loading the entire tidyverse.
* ggplot2 uses a *grammar of graphics* to break plots into building blocks that have intuitive syntax, making it easy to create relatively complex and aesthetically pleasing plots with relatively simple and readable code.
* ggplot2 is designed to work exclusively with tidy data (rows are observations and columns are variables).
* We recommend you keep a copy of the ggplot2 cheat sheet handy.

### 2.1.2 Graph Components

**Key points**

* Plots in ggplot2 consist of 3 main components:  
  + Data: The dataset being summarized  
  + Geometry: The type of plot (scatterplot, boxplot, barplot, histogram, qqplot, smooth density, etc.)  
  + Aesthetic mapping: Variables mapped to visual cues, such as x-axis and y-axis values and color  
* There are additional components:  
  + Scale
  + Labels, Title, Legend
  + Theme/Style  

**Code**
```{r}
library(dslabs)
data(murders)
```

### 2.1.3 Creating a new plot

**Key points**

* You can associate a dataset x with a ggplot object with any of the 3 commands:  
  + ggplot(data = x)  
  + ggplot(x)  
  + x %>% ggplot()  
* You can assign a ggplot object to a variable. If the object is not assigned to a variable, it will automatically be displayed.
* You can display a ggplot object assigned to a variable by printing that variable.
  
**Code**
```{r}
ggplot(data = murders)
```

```{r}
murders %>% ggplot
```

```{r}
p <- ggplot(data = murders)
class(p)
print(p)    # this is equivalent to simply typing p
```


## 2.2 Customizing Plots

```{r}
library(tidyverse)
library(dslabs)
data(murders)
p <- ggplot(data = murders)
```

### 2.2.1 Layers

**Key points**

* In ggplot2, graphs are created by adding layers to the ggplot object:
DATA %>% ggplot() + LAYER_1 + LAYER_2 + ... + LAYER_N
* The geometry layer defines the plot type and takes the format geom_X where X is the plot type.
* Aesthetic mappings describe how properties of the data connect with features of the graph (axis position, color, size, etc.) Define aesthetic mappings with the aes function.
* aes uses variable names from the object component (for example, total rather than murders$total).
* geom_point creates a scatterplot and requires x and y aesthetic mappings. 
* geom_text and geom_label add text to a scatterplot and require x, y, and label aesthetic mappings.
* To determine which aesthetic mappings are required for a geometry, read the help file for that geometry.
* You can add layers with different aesthetic mappings to the same graph.


**Code: Adding layers to a plot**

```{r}
murders %>% ggplot() +
    geom_point(aes(x = population/10^6, y = total))
```

```{r}
# Add points layer to predefined ggplot object
p <- ggplot(data = murders)
p + geom_point(aes(population/10^6, total))
```

```{r}
# Add text layer to scatterplot
p + geom_point(aes(population/10^6, total)) +
    geom_text(aes(population/10^6, total, label = abb))
```


### 2.2.2 Tinkering

**Key points**

* You can modify arguments to geometry functions other than aes and the data. Additional arguments can be found in the documentation for each geometry.  
* These arguments are not aesthetic mappings: they affect all data points the same way.  
* Global aesthetic mappings apply to all geometries and can be defined when you initially call ggplot. All the geometries added as layers will default to this mapping. Local aesthetic mappings add additional information or override the default mappings.

**Code**

```{r}
# change the size of the points
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb))
```

```{r}
# move text labels slightly to the right
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb), nudge_x = 1)
```

```{r}
# simplify code by adding global aesthetic
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
p + geom_point(size = 3) +
    geom_text(nudge_x = 1.5)
```

```{r}
# local aesthetics override global aesthetics
p + geom_point(size = 3) +
    geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```


### 2.2.3 Scales, Labels, and Colors

**Key points**

* Convert the x-axis to log scale with scale_x_continuous(trans = "log10") or scale_x_log10. Similar functions exist for the y-axis.  
* Add axis titles with xlab and ylab functions. Add a plot title with the ggtitle function.  
* Add a color mapping that colors points by a variable by defining the col argument within aes. To color all points the same way, define col outside of aes.
* Add a line with the geom_abline geometry. geom_abline takes arguments slope (default = 1) and intercept (default = 0). Change the color with col or color and line type with lty.
* Placing the line layer after the point layer will overlay the line on top of the points. To overlay points on the line, place the line layer before the point layer.
* There are many additional ways to tweak your graph that can be found in the ggplot2 documentation, cheat sheet, or on the internet. For example, you can change the legend title with scale_color_discrete.

**Code: Log-scale the x- and y-axis**

```{r}
# define p
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))

# log base 10 scale the x-axis and y-axis
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.05) +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10")

# efficient log scaling of the axes
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10()
```

**Code: Add labels and title**

```{r}
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")
```

**Code: Change color of the points**

```{r}
# redefine p to be everything except the points layer
p <- murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")

# make all points blue
p + geom_point(size = 3, color = "blue")

# color points by region
p + geom_point(aes(col = region), size = 3)
```

**Code: Add a line with average murder rate**

```{r}
# define average murder rate
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    pull(rate)

# basic line with average murder rate for the country
p + geom_point(aes(col = region), size = 3) +
    geom_abline(intercept = log10(r))    # slope is default of 1

# change line to dashed and dark grey, line under points
p <- p + geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3)
p
```

**Code: Change legend title**

```{r}
# capitalize legend title
p <- p + scale_color_discrete(name = "Region")
p
```


### 2.2.4 Add-on Packages

**Key points**

* The style of a ggplot graph can be changed using the theme function.
* The **ggthemes** package adds additional themes.
The **ggrepel** package includes a geometry that repels text labels, ensuring they do not overlap with each other: geom_text_repel.

**Code: Adding themes**

```{r}
# theme used for graphs in the textbook and course
library(dslabs)
ds_theme_set()

# themes from ggthemes
library(ggthemes)
p + theme_economist()    # style of the Economist magazine
p + theme_fivethirtyeight()    # style of the FiveThirtyEight website
```

**Code: Putting it all together to assemble the plot**

```{r}
# load libraries
library(tidyverse)
library(ggrepel)
library(ggthemes)
library(dslabs)
data(murders)

# define the intercept
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    .$rate

# make the plot, combining all elements
murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3) +
    geom_text_repel() +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010") +
    scale_color_discrete(name = "Region") +
    theme_economist()
```


### 2.2.5 Other Examples

**Key points**

* *geom_histogram* creates a histogram. Use the *binwidth* argument to change the width of bins, the fill argument to change the bar fill color, and the col argument to change bar outline color.
* *geom_density* creates smooth density plots. Change the fill color of the plot with the *fill* argument.
* *geom_qq* creates a quantile-quantile plot. This geometry requires the *sample* argument. By default, the data are compared to a standard normal distribution with a mean of 0 and standard deviation of 1. This can be changed with the *dparams* argument, or the sample data can be scaled.
* Plots can be arranged adjacent to each other using the *grid.arrange* function from the *gridExtra* package. First, create the plots and save them to objects (p1, p2, ...). Then pass the plot objects to *grid.arrange.*

**Code: Histograms in ggplot2**

```{r}
# load heights data
library(tidyverse)
library(dslabs)
data(heights)

# define p
p <- heights %>%
    filter(sex == "Male") %>%
    ggplot(aes(x = height))

# basic histograms
p + geom_histogram()
p + geom_histogram(binwidth = 1)

# histogram with blue fill, black outline, labels and title
p + geom_histogram(binwidth = 1, fill = "blue", col = "black") +
    xlab("Male heights in inches") +
    ggtitle("Histogram")
```

**Code: Smooth density plots in ggplot2**

```{r}
p + geom_density()
p + geom_density(fill = "blue")
```

**Code: Quantile-quantile plots in ggplot2**

```{r}
# basic QQ-plot
p <- heights %>% filter(sex == "Male") %>%
    ggplot(aes(sample = height))
p + geom_qq()

# QQ-plot against a normal distribution with same mean/sd as data
params <- heights %>%
    filter(sex == "Male") %>%
    summarize(mean = mean(height), sd = sd(height))
p + geom_qq(dparams = params) +
    geom_abline()

# QQ-plot of scaled data against the standard normal distribution
heights %>%
    ggplot(aes(sample = scale(height))) +
    geom_qq() +
    geom_abline()
```

**Code: Grids of plots with the grid.extra package**

```{r}
# define plots p1, p2, p3
p <- heights %>% filter(sex == "Male") %>% ggplot(aes(x = height))
p1 <- p + geom_histogram(binwidth = 1, fill = "blue", col = "black")
p2 <- p + geom_histogram(binwidth = 2, fill = "blue", col = "black")
p3 <- p + geom_histogram(binwidth = 3, fill = "blue", col = "black")

# arrange plots next to each other in 1 row, 3 columns
library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 3)
```

### 2.2.6 Assessment: Introduction to ggplot2
...

