---
title: "Section2: Introduction to ggplot2"
output: html_notebook
---

# Basics of ggplot2

## ggplot2

**Key points and notes**

* Throughout the series, we will create plots with the ggplot2 package. ggplot2 is part of the tidyverse, which you can load with library(tidyverse).
* Note that you can also load ggplot2 alone using the command library(ggplot2), instead of loading the entire tidyverse.
* ggplot2 uses a grammar of graphics to break plots into building blocks that have intuitive syntax, making it easy to create relatively complex and aesthetically pleasing plots with relatively simple and readable code.
* ggplot2 is designed to work exclusively with tidy data (rows are observations and columns are variables).
* We recommend you keep a copy of the ggplot2 cheat sheet handy.

```{r}
library(tidyverse)
```

## Graph Components

**Key points**

* Plots in ggplot2 consist of 3 main components:  
  + Data: The dataset being summarized  
  + Geometry: The type of plot (scatterplot, boxplot, barplot, histogram, qqplot, smooth density, etc.)  
  + Aesthetic mapping: Variables mapped to visual cues, such as x-axis and y-axis values and color  
* There are additional components:  
  + Scale
  + Labels, Title, Legend
  + Theme/Style  

**Code**
```{r}
library(dslabs)
data(murders)
```

## Creating a new plot

**Key points**

* You can associate a dataset x with a ggplot object with any of the 3 commands:  
  + ggplot(data = x)  
  + ggplot(x)  
  + x %>% ggplot()  
* You can assign a ggplot object to a variable. If the object is not assigned to a variable, it will automatically be displayed.
* You can display a ggplot object assigned to a variable by printing that variable.
  
**Code**
```{r}
ggplot(data = murders)
```

```{r}
murders %>% ggplot
```

```{r}
p <- ggplot(data = murders)
class(p)
print(p)    # this is equivalent to simply typing p
```

