---
title: 'Data Science: Wrangling - HarvardX: PH125.6x'
author: 'Luiz Cunha'
date: '2019-08-13'
output: html_notebook
---

# Section 3: String Processing

## Overview

In the **String Processing** section, we use case studies that help demonstrate how string processing is a powerful tool useful for overcoming many data wrangling challenges. You will see how the original raw data was processed to create the data frames we have used in courses throughout this series.

This section is divided into three parts. There are comprehension checks at the end of each of the three parts.  
After completing the **String Processing** section, you will be able to:

* **Remove** unwanted characters from text.
* **Extract numeric** values from text.
* **Find** and **replace** characters.
* **Extract specific** parts of strings.
* **Convert** free form text into more uniform formats.
* **Split strings** into multiple values.
* Use **regular expressions (regex)** to process strings.


## 3.1 String Processing, Part 1

### 3.1.1 String Parsing

**Key points**

* The most common tasks in string processing include:
  + extracting numbers from strings
  + removing unwanted characters from text
  + finding and replacing characters
  + extracting specific parts of strings
  + converting free form text to more uniform formats
  + splitting strings into multiple values
* The **stringr** package in the **tidyverse** contains string processing functions that follow a similar naming format (str_functionname) and are compatible with the pipe.

**Code**
```{r}
# read in raw murders data from Wikipedia
url <- "https://en.wikipedia.org/w/index.php?title=Gun_violence_in_the_United_States_by_state&direction=prev&oldid=810166167"
murders_raw <- read_html(url) %>% 
  html_nodes("table") %>% 
  html_table() %>%
  .[[2]] %>%
  setNames(c("state", "population", "total", "murder_rate"))
# inspect data and column classes
head(murders_raw)
class(murders_raw$population)
class(murders_raw$total)
```

### 3.1.2 Defining Strings: Single and Double Quotes and How to Escape

**Key points**

* Define a string by surrounding text with either single quotes or double quotes.
* To include a single quote inside a string, use double quotes on the outside. To include a double quote inside a string, use single quotes on the outside.
* The cat function displays a string as it is represented inside R.
* To include a double quote inside of a string surrounded by double quotes, use the backslash (\) to escape the double quote. Escape a single quote to include it inside of a string defined by single quotes.
* We will see additional uses of the escape later.

**Code**
```{r eval=FALSE}
s <- "Hello!"    # double quotes define a string
s <- 'Hello!'    # single quotes define a string
s <- `Hello`    # backquotes do not
s <- "10""    # error - unclosed quotes
s <- '10"'    # correct
# cat shows what the string actually looks like inside R
cat(s)
s <- "5'"
cat(s)
# to include both single and double quotes in string, escape with \
s <- '5'10"'    # error
s <- "5'10""    # error
s <- '5\'10"'    # correct
cat(s)
s <- "5'10\""    # correct
cat(s)
```


### 3.1.3 stringr Package

**Key points**
* The main types of string processing tasks are detecting, locating, extracting and replacing elements of strings.
* The **stringr** package from **tidyverse** includes a variety of string processing functions that begin with str_ and take the string as the first argument, which makes them compatible with the pipe.

**Code**

# murders_raw defined in web scraping video
# direct conversion to numeric fails because of commas
```{r}
murders_raw$population[1:3]
as.numeric(murders_raw$population[1:3])
library(tidyverse)    # includes stringr
```


### 3.1.4 Case Study 1: US Murders Data

**Key points**

* Use the *str_detect* function to determine whether a string contains a pattern.
* Use the *str_replace_all* function to replace all instances of one pattern with another pattern. To remove a pattern, replace with the empty string ("").
* The *parse_number* function removes punctuation from strings and converts them to numeric.
mutate_at performs the same transformation on the specified column numbers.

**Code**
```{r}
# murders_raw was defined in the web scraping section
# detect whether there are commas
commas <- function(x) any(str_detect(x, ","))
murders_raw %>% summarize_all(funs(commas))
# replace commas with the empty string and convert to numeric
test_1 <- str_replace_all(murders_raw$population, ",", "")
test_1 <- as.numeric(test_1)

# parse_number also removes commas and converts to numeric
test_2 <- parse_number(murders_raw$population)
identical(test_1, test_2)
murders_new <- murders_raw %>% mutate_at(2:3, parse_number)
murders_new %>% head
```


## 3.2 String Processing, Part 2

### 3.2.1 Case Study 2: Reported Heights

**Key points**

* In the raw heights data, many students did not report their height as the number of inches as requested. There are many entries with real height information but in the wrong format, which we can extract with string processing. 
* When there are both text and numeric entries in a column, the column will be a character vector. Converting this column to numeric will result in NAs for some entries.
* To correct problematic entries, look for patterns that are shared across large numbers of entries, then define rules that identify those patterns and use these rules to write string processing tasks.
* Use *suppressWarnings* to hide warning messages for a function.

**Code**
```{r}
# load raw heights data and inspect
library(dslabs)
data(reported_heights)
class(reported_heights$height)

# convert to numeric, inspect, count NAs
x <- as.numeric(reported_heights$height)
head(x)
sum(is.na(x))

# keep only entries that result in NAs
reported_heights %>% mutate(new_height = as.numeric(height)) %>%
  filter(is.na(new_height)) %>% 
  head(n=10)

# calculate cutoffs that cover 99.999% of human population.
alpha <- 1/10^6
qnorm(1-alpha/2, 69.1, 2.9)
qnorm(alpha/2, 63.7, 2.7)

# keep only entries that either result in NAs or are outside the plausible range of heights
not_inches <- function(x, smallest = 50, tallest = 84){
  inches <- suppressWarnings(as.numeric(x))
  ind <- is.na(inches) | inches < smallest | inches > tallest
  ind
}

# number of problematic entries
problems <- reported_heights %>% 
  filter(not_inches(height)) %>%
  .$height
length(problems)

# 10 examples of x'y or x'y" or x'y\''
pattern <- "^\\d\\s*'\\s*\\d{1,2}\\.*\\d*'*\"*$"
str_subset(problems, pattern) %>% head(n=10) %>% cat

# 10 examples of x.y or x,y
pattern <- "^[4-6]\\s*[\\.|,]\\s*([0-9]|10|11)$"
str_subset(problems, pattern) %>% head(n=10) %>% cat

# 10 examples of entries in cm rather than inches
ind <- which(between(suppressWarnings(as.numeric(problems))/2.54, 54, 81) )
ind <- ind[!is.na(ind)]
problems[ind] %>% head(n=10) %>% cat
```

### 3.2.2 Regex

**Key points**

* A regular expression (regex) is a way to describe a specific pattern of characters of text. A set of rules has been designed to do this specifically and efficiently.
* **stringr** functions can take a regex as a pattern.
* *str_detect* indicates whether a pattern is present in a string.
* The main difference between a regex and a regular string is that a regex can include special characters.
* The | symbol inside a regex means "or".
* Use '\\d' to represent digits. The backlash is used to distinguish it from the character 'd'. In R, you must use two backslashes for digits in regular expressions; in some other languages, you will only use one backslash for regex special characters.
* *str_view* highlights the first occurrence of a pattern, and the *str_view_all* function highlights all occurrences of the pattern.

**Code**
```{r}
# load stringr through tidyverse
library(tidyverse)

# detect whether a comma is present
pattern <- ","
str_detect(murders_raw$total, pattern) 

# show the subset of strings including "cm"
str_subset(reported_heights$height, "cm")

# use the "or" symbol inside a regex (|)
yes <- c("180 cm", "70 inches")
no <- c("180", "70''")
s <- c(yes, no)
str_detect(s, "cm") | str_detect(s, "inches")
str_detect(s, "cm|inches")

# highlight the first occurrence of a pattern
str_view(s, pattern)

# highlight all instances of a pattern
str_view_all(s, pattern)
```

### 3.2.3 Character Classes, Anchors and Quantifiers

**Key points**

* Define strings to test your regular expressions, including some elements that match and some that do not. This allows you to check for the two types of errors: failing to match and matching incorrectly.
* Square brackets define character classes: groups of characters that count as matching the pattern. You can use ranges to define character classes, such as [0-9] for digits and [a-zA-Z] for all letters.
* Anchors define patterns that must start or end at specific places. ^ and $ represent the beginning and end of the string respectively.
* Curly braces are quantifiers that state how many times a certain character can be repeated in the pattern. \\d{1,2} matches exactly 1 or 2 consecutive digits.

**Code**
```{r}
# s was defined in the previous video
yes <- c("5", "6", "5'10", "5 feet", "4'11")
no <- c("", ".", "Five", "six")
s <- c(yes, no)
pattern <- "\\d"

# [56] means 5 or 6
str_view(s, "[56]")

# [4-7] means 4, 5, 6 or 7
yes <- as.character(4:7)
no <- as.character(1:3)
s <- c(yes, no)
str_detect(s, "[4-7]")

# ^ means start of string, $ means end of string
pattern <- "^\\d$"
yes <- c("1", "5", "9")
no <- c("12", "123", " 1", "a4", "b")
s <- c(yes, no)
str_view(s, pattern)

# curly braces define quantifiers: 1 or 2 digits 
pattern <- "^\\d{1,2}$"
yes <- c("1", "5", "9", "12")
no <- c("123", "a4", "b")
str_view(c(yes, no), pattern)

# combining character class, anchors and quantifier
pattern <- "^[4-7]'\\d{1,2}\"$"
yes <- c("5'7\"", "6'2\"",  "5'12\"")
no <- c("6,2\"", "6.2\"","I am 5'11\"", "3'2\"", "64")
str_detect(yes, pattern)
str_detect(no, pattern)
```

### 3.2.4 Search and Replace Regex

**Key points**

* *str_replace* replaces the first instance of the detected pattern with a specified string.
* Spaces are characters and R does not ignore them. Spaces are specified by the special character \\s.
* Additional quantifiers include **\***, **+** and **?**. 
  + \* means 0 or more instances of the previous character. 
  + ? means 0 or 1 instances. 
  + \+ means 1 or more instances.
* Before removing characters from strings with functions like *str_replace* and *str_replace_all*, consider whether that replacement would have unintended effects.

**Code**

The problems object is defined in the reported heights case study introduction video.
```{r}
# number of entries matching our desired pattern
pattern <- "^[4-7]'\\d{1,2}\"$"
sum(str_detect(problems, pattern))

# inspect examples of entries with problems
problems[c(2, 10, 11, 12, 15)] %>% str_view(pattern)
str_subset(problems, "inches")
str_subset(problems, "''")

# replace or remove feet/inches words before matching
pattern <- "^[4-7]'\\d{1,2}$"
problems %>% 
  str_replace("feet|ft|foot", "'") %>% # replace feet, ft, foot with ' 
  str_replace("inches|in|''|\"", "") %>% # remove all inches symbols
  str_detect(pattern) %>% 
  sum()

# R does not ignore whitespace
identical("Hi", "Hi ")

# \\s represents whitespace
pattern_2 <- "^[4-7]'\\s\\d{1,2}\"$"
str_subset(problems, pattern_2)

# * means 0 or more instances of a character
yes <- c("AB", "A1B", "A11B", "A111B", "A1111B")
no <- c("A2B", "A21B")
str_detect(yes, "A1*B")
str_detect(no, "A1*B")

# test how *, ? and + differ
data.frame(string = c("AB", "A1B", "A11B", "A111B", "A1111B"),
           none_or_more = str_detect(yes, "A1*B"),
           nore_or_once = str_detect(yes, "A1?B"),
           once_or_more = str_detect(yes, "A1+B"))

# update pattern by adding optional spaces before and after feet symbol
pattern <- "^[4-7]\\s*'\\s*\\d{1,2}$"
problems %>% 
  str_replace("feet|ft|foot", "'") %>% # replace feet, ft, foot with ' 
  str_replace("inches|in|''|\"", "") %>% # remove all inches symbols
  str_detect(pattern) %>% 
  sum()
```

  
### 3.2.5 Groups with Regex

**Key Points**

* Groups are defined using parentheses.
* Once we define groups, we can use the function *str_match* to extract the values these groups define. *str_extract* extracts only strings that match a pattern, not the values defined by groups.
* You can refer to the ith group with \\i. For example, refer to the value in the second group with \\2.

**Code**
```{r}
# define regex with and without groups
pattern_without_groups <- "^[4-7],\\d*$"
pattern_with_groups <-  "^([4-7]),(\\d*)$"

# create examples
yes <- c("5,9", "5,11", "6,", "6,1")
no <- c("5'9", ",", "2,8", "6.1.1")
s <- c(yes, no)

# demonstrate the effect of groups
str_detect(s, pattern_without_groups)
str_detect(s, pattern_with_groups)

# demonstrate difference between str_match and str_extract
str_match(s, pattern_with_groups)
str_extract(s, pattern_with_groups)

# improve the pattern to recognize more events
pattern_with_groups <-  "^([4-7]),(\\d*)$"
yes <- c("5,9", "5,11", "6,", "6,1")
no <- c("5'9", ",", "2,8", "6.1.1")
s <- c(yes, no)
str_replace(s, pattern_with_groups, "\\1'\\2")

#final pattern
pattern_with_groups <-"^([4-7])\\s*[,\\.\\s+]\\s*(\\d*)$"

# combine stringr commands with the pipe
str_subset(problems, pattern_with_groups) %>% head
str_subset(problems, pattern_with_groups) %>% 
  str_replace(pattern_with_groups, "\\1'\\2") %>% head
```


### 3.2.6 Testing and Improving

**Key points**

* Wrangling with regular expressions is often an iterative process of testing the approach, looking for problematic entries, and improving the patterns.
* Use the pipe to connect **stringr** functions.
* It may not be worth writing code to correct every unique problem in the data, but string processing techniques are flexible enough for most needs.

**Code**
```{r}
# function to detect entries with problems
not_inches_or_cm <- function(x, smallest = 50, tallest = 84){
    inches <- suppressWarnings(as.numeric(x))
    ind <- !is.na(inches) &
        ((inches >= smallest & inches <= tallest) |
             (inches/2.54 >= smallest & inches/2.54 <= tallest))
    !ind
}

# identify entries with problems
problems <- reported_heights %>% 
  filter(not_inches_or_cm(height)) %>%
  .$height
length(problems)
converted <- problems %>% 
  str_replace("feet|foot|ft", "'") %>% #convert feet symbols to '
  str_replace("inches|in|''|\"", "") %>%  #remove inches symbols
  str_replace("^([4-7])\\s*[,\\.\\s+]\\s*(\\d*)$", "\\1'\\2") ##change format

# find proportion of entries that fit the pattern after reformatting
pattern <- "^[4-7]\\s*'\\s*\\d{1,2}$"
index <- str_detect(converted, pattern)
mean(index)

converted[!index]    # show problems
```


### 3.2.7 Assessment
...

## 3.3 String Processing, Part 3

### 3.3.1 Separate with Regex

**Key Point**

* The *extract* function behaves similarly to the separate function but allows extraction of groups from regular expressions.

**Code**
```{r}
# first example - normally formatted heights
s <- c("5'10", "6'1")
tab <- data.frame(x = s)

# the separate and extract functions behave similarly
tab %>% separate(x, c("feet", "inches"), sep = "'")
tab %>% extract(x, c("feet", "inches"), regex = "(\\d)'(\\d{1,2})")

# second example - some heights with unusual formats
s <- c("5'10", "6'1\"","5'8inches")
tab <- data.frame(x = s)

# separate fails because it leaves in extra characters, but extract keeps only the digits because of regex groups
tab %>% separate(x, c("feet","inches"), sep = "'", fill = "right")
tab %>% extract(x, c("feet", "inches"), regex = "(\\d)'(\\d{1,2})")
```


### 3.3.2 Using Groups and Quantifiers

Four clear patterns of entries have arisen along with some other minor problems:

1. Many students measuring exactly 5 or 6 feet did not enter any inches. For example, 6' - our pattern requires that inches be included.
2. Some students measuring exactly 5 or 6 feet entered just that number.
3. Some of the inches were entered with decimal points. For example 5'7.5''. Our pattern only looks for two digits.
4. Some entires have spaces at the end, for example 5 ' 9.
5. Some entries are in meters and some of these use European decimals: 1.6, 1,7.
6. Two students added cm.
7. One student spelled out the numbers: Five foot eight inches.

It is not necessarily clear that it is worth writing code to handle all these cases since they might be rare enough. However, some give us an opportunity to learn some more regex techniques so we will build a fix.

**Case 1**

For case 1, if we add a '0 to, for example, convert all 6 to 6'0, then our pattern will match. This can be done using groups using the following code:
```{r}
yes <- c("5", "6", "5")
no <- c("5'", "5''", "5'4")
s <- c(yes, no)
str_replace(s, "^([4-7])$", "\\1'0")
```

The pattern says it has to start (^), be followed with a digit between 4 and 7, and then end there ($). The parenthesis defines the group that we pass as \\1 to the replace regex.

**Cases 2 and 4**
We can adapt this code slightly to handle case 2 as well which covers the entry 5'. Note that the 5' is left untouched by the code above. This is because the extra ' makes the pattern not match since we have to end with a 5 or 6. To handle case 2, we want to permit the 5 or 6 to be followed by no or one symbol for feet. So we can simply add '{0,1} after the ' to do this. We can also use the none or once special character ?. As we saw previously, this is different from * which is none or more. We now see that this code also handles the fourth case as well:
```{r}
str_replace(s, "^([56])'?$", "\\1'0")
```

Note that here we only permit 5 and 6 but not 4 and 7. This is because heights of exactly 5 and exactly 6 feet tall are quite common, so we assume those that typed 5 or 6 really meant either 60 or 72 inches. However, heights of exactly 4 or exactly 7 feet tall are so rare that, although we accept 84 as a valid entry, we assume that a 7 was entered in error.

**Case 3**

We can use quantifiers to deal with  case 3. These entries are not matched because the inches include decimals and our pattern does not permit this. We need allow the second group to include decimals and not just digits. This means we must permit zero or one period . followed by zero or more digits. So we will use both ? and *. Also remember that for this particular case, the period needs to be escaped since it is a special character (it means any character except a line break).

So we can adapt our pattern, currently ^[4-7]\\s*'\\s*\\d{1,2}$, to permit a decimal at the end:
```{r}
pattern <- "^[4-7]\\s*'\\s*(\\d+\\.?\\d*)$"
```

**Case 5**

Case 5, meters using commas, we can approach similarly to how we converted the x.y to x'y. A difference is that we require that the first digit is 1 or 2:
```{r}
yes <- c("1,7", "1, 8", "2, " )
no <- c("5,8", "5,3,2", "1.7")
s <- c(yes, no)
str_replace(s, "^([12])\\s*,\\s*(\\d*)$", "\\1\\.\\2")
```

We will later check if the entries are meters using their numeric values.

**Trimming**

In general, spaces at the start or end of the string are uninformative. These can be particularly deceptive because sometimes they can be hard to see:
```{r}
s <- "Hi "
cat(s)
identical(s, "Hi")
```
This is a general enough problem that there is a function dedicated to removing them: str_trim.
```{r}
str_trim("5 ' 9 ")
```

**To upper and to lower case**

One of the entries writes out numbers as words: Five foot eight inches. Although not efficient, we could add 12 extra str_replace to convert zero to 0, one to 1, and so on. To avoid having to write two separate operations for Zero and zero, One and one, etc., we can use the str_to_lower function to make all words lower case first:
```{r}
s <- c("Five feet eight inches")
str_to_lower(s)
```

**Putting it into a function**

We are now ready to define a procedure that handles converting all the problematic cases.

We can now put all this together into a function that takes a string vector and tries to convert as many strings as possible to a single format. Below is a function that puts together the previous code replacements:
```{r}
convert_format <- function(s){
  s %>%
    str_replace("feet|foot|ft", "'") %>% #convert feet symbols to '
    str_replace_all("inches|in|''|\"|cm|and", "") %>%  #remove inches and other symbols
    str_replace("^([4-7])\\s*[,\\.\\s+]\\s*(\\d*)$", "\\1'\\2") %>% #change x.y, x,y x y
    str_replace("^([56])'?$", "\\1'0") %>% #add 0 when to 5 or 6
    str_replace("^([12])\\s*,\\s*(\\d*)$", "\\1\\.\\2") %>% #change european decimal
    str_trim() #remove extra space
}
```

We can also write a function that converts words to numbers:
```{r}
words_to_numbers <- function(s){
  str_to_lower(s) %>%  
    str_replace_all("zero", "0") %>%
    str_replace_all("one", "1") %>%
    str_replace_all("two", "2") %>%
    str_replace_all("three", "3") %>%
    str_replace_all("four", "4") %>%
    str_replace_all("five", "5") %>%
    str_replace_all("six", "6") %>%
    str_replace_all("seven", "7") %>%
    str_replace_all("eight", "8") %>%
    str_replace_all("nine", "9") %>%
    str_replace_all("ten", "10") %>%
    str_replace_all("eleven", "11")
}
```

Now we can see which problematic entries remain:
```{r}
converted <- problems %>% words_to_numbers %>% convert_format
remaining_problems <- converted[not_inches_or_cm(converted)]
pattern <- "^[4-7]\\s*'\\s*\\d+\\.?\\d*$"
index <- str_detect(remaining_problems, pattern)
remaining_problems[!index]
```


### 3.3.3 Putting it All Together

We are now ready to put everything we've done so far together and wrangle our reported heights data as we try to recover as many heights as possible. The code is complex but we will break it down into parts.

We start by cleaning up the height column so that the heights are closer to a feet'inches format. We added an original heights column so we can compare before and after.

Let's start by writing a function that cleans up strings so that all the feet and inches formats use the same x'y format when appropriate.
```{r}
pattern <- "^([4-7])\\s*'\\s*(\\d+\\.?\\d*)$"

smallest <- 50
tallest <- 84
new_heights <- reported_heights %>% 
  mutate(original = height, 
         height = words_to_numbers(height) %>% convert_format()) %>%
  extract(height, c("feet", "inches"), regex = pattern, remove = FALSE) %>% 
  mutate_at(c("height", "feet", "inches"), as.numeric) %>%
  mutate(guess = 12*feet + inches) %>%
  mutate(height = case_when(
    !is.na(height) & between(height, smallest, tallest) ~ height, #inches 
    !is.na(height) & between(height/2.54, smallest, tallest) ~ height/2.54, #centimeters
    !is.na(height) & between(height*100/2.54, smallest, tallest) ~ height*100/2.54, #meters
    !is.na(guess) & inches < 12 & between(guess, smallest, tallest) ~ guess, #feet'inches
    TRUE ~ as.numeric(NA))) %>%
  select(-guess)
```

We can check all the entries we converted using the following code:
```{r}
new_heights %>%
  filter(not_inches(original)) %>%
  select(original, height) %>% 
  arrange(height) %>%
  View()
```

Let's take a look at the shortest students in our dataset using the following code:
```{r}
new_heights %>% arrange(height) %>% head(n=7)
```

We see heights of 53, 54, and 55. In the original heights column, we also have 51 and 52. These short heights are very rare and it is likely that the students actually meant 5'1, 5'2, 5'3, 5'4, and 5'5. But because we are not completely sure, we will leave them as reported.


### 3.3.4 String Splitting

**Key Points**

* The function *str_split* splits a string into a character vector on a delimiter (such as a comma, space or underscore). By default, *str_split* generates a list with one element for each original string. Use the function argument *simplify=TRUE* to have *str_split* return a matrix instead.
* The *map* function from the **purrr** package applies the same function to each element of a list. To extract the ith entry of each element x, use *map(x, i)*.
* *map* always returns a list. Use *map_chr* to return a character vector and *map_int* to return an integer.

**Code**
```{r}
# read raw murders data line by line
filename <- system.file("extdata/murders.csv", package = "dslabs")
lines <- readLines(filename)
lines %>% head()

# split at commas with str_split function, remove row of column names
x <- str_split(lines, ",") 
x %>% head()
col_names <- x[[1]]
x <- x[-1]

# extract first element of each list entry
library(purrr)
map(x, function(y) y[1]) %>% head()
map(x, 1) %>% head()

# extract columns 1-5 as characters, then convert to proper format
dat <- data.frame(map_chr(x, 1),  
                  map_chr(x, 2),
                  map_chr(x, 3),
                  map_chr(x, 4),
                  map_chr(x, 5)) %>%
  mutate_all(parse_guess) %>%
  setNames(col_names)
dat %>% head

# more efficient code for the same thing
dat <- x %>%
  transpose() %>%
  map( ~ parse_guess(unlist(.))) %>%
  setNames(col_names) %>% 
  as.data.frame() 

# the simplify argument makes str_split return a matrix instead of a list
x <- str_split(lines, ",", simplify = TRUE) 
col_names <- x[1,]
x <- x[-1,]
x %>% as_data_frame() %>%
  setNames(col_names) %>%
  mutate_all(parse_guess)
```


### 3.3.5 Case Study: extracting a table from a PDF

One of the datasets provided in dslabs shows scientific funding rates by gender in the Netherlands:
```{r}
library(dslabs)
data("research_funding_rates")
research_funding_rates 
```

The data come from a paper published in the prestigious journal PNAS. However, the data are not provided in a spreadsheet; they are in a table in a PDF document. We could extract the numbers by hand, but this could lead to human error. Instead we can try to wrangle the data using R.

**Downloading the data**

We start by downloading the PDF document then importing it into R using the following code:
```{r}
library("pdftools")
temp_file <- tempfile()
url <- "http://www.pnas.org/content/suppl/2015/09/16/1510159112.DCSupplemental/pnas.201510159SI.pdf"
download.file(url, temp_file)
txt <- pdf_text(temp_file)
file.remove(temp_file)
```

If we examine the object text we notice that it is a character vector with an entry for each page. So we keep the page we want using the following code:
```{r}
raw_data_research_funding_rates <- txt[2]
```

The steps above can actually be skipped because we include the raw data in the dslabs package as well:
```{r}
data("raw_data_research_funding_rates")
```

**Looking at the download**

Examining this object,
```{r}
raw_data_research_funding_rates %>% head
```

we see that it is a long string. Each line on the page, including the table rows, is separated by the symbol for newline: \n.

We can therefore can create a list with the lines of the text as elements:
```{r}
tab <- str_split(raw_data_research_funding_rates, "\n")
```

Because we start off with just one element in the string, we end up with a list with just one entry:
```{r}
tab <- tab[[1]]
```

By examining this object,
```{r}
tab %>% head
```

we see that the information for the column names is the third and forth entires:
```{r}
the_names_1 <- tab[3]
the_names_2 <- tab[4]
```

In the table, the column information is spread across two lines. We want to create one vector with one name for each column. We can do this using some of the functions we have just learned.

**Extracting the table data**

Let's start with the first line:

```{r}
the_names_1
```

We want to remove the leading space and everything following the comma. We can use regex for the latter. Then we can obtain the elements by splitting using the space. We want to split only when there are 2 or more spaces to avoid splitting success rate. So we use the regex \\s{2,} as follows:

```{r}
the_names_1 <- the_names_1 %>%
  str_trim() %>%
  str_replace_all(",\\s.", "") %>%
  str_split("\\s{2,}", simplify = TRUE)
the_names_1
```

Now let's look at the second line:
```{r}
the_names_2
```

Here we want to trim the leading space and then split by space as we did for the first line:

```{r}
the_names_2 <- the_names_2 %>%
  str_trim() %>%
  str_split("\\s+", simplify = TRUE)
the_names_2
```

Now we can join these to generate one name for each column:
```{r}
tmp_names <- str_c(rep(the_names_1, each = 3), the_names_2[-1], sep = "_")
the_names <- c(the_names_2[1], tmp_names) %>%
  str_to_lower() %>%
  str_replace_all("\\s", "_")
the_names
```

Now we are ready to get the actual data. By examining the tab object, we notice that the information is in lines 6 through 14. We can use str_split again to achieve our goal:

```{r}
new_research_funding_rates <- tab[6:14] %>%
  str_trim %>%
  str_split("\\s{2,}", simplify = TRUE) %>%
  data.frame(stringsAsFactors = FALSE) %>%
  setNames(the_names) %>%
  mutate_at(-1, parse_number)
new_research_funding_rates %>% head()
```

We can see that the objects are identical:
```{r}
identical(research_funding_rates, new_research_funding_rates)
```

### 3.3.6 Recoding

**Key points**

Change long factor names with the recode function from the tidyverse. 
Other similar functions include recode_factor and fct_recoder in the forcats package in the tidyverse. The same result could be obtained using the case_when function, but recode is more efficient to write.

**Code**
```{r}
# life expectancy time series for Caribbean countries
library(dslabs)
data("gapminder")
gapminder %>% 
  filter(region=="Caribbean") %>%
  ggplot(aes(year, life_expectancy, color = country)) +
  geom_line()

# display long country names
gapminder %>% 
  filter(region=="Caribbean") %>%
  filter(str_length(country) >= 12) %>%
  distinct(country) 

# recode long country names and remake plot
gapminder %>% filter(region=="Caribbean") %>%
  mutate(country = recode(country, 
                          'Antigua and Barbuda'="Barbuda",
                          'Dominican Republic' = "DR",
                          'St. Vincent and the Grenadines' = "St. Vincent",
                          'Trinidad and Tobago' = "Trinidad")) %>%
  ggplot(aes(year, life_expectancy, color = country)) +
  geom_line()
```


### 3.3.7 Assessment Part 1
...


### 3.3.8 Assessment Part 2

Import raw Brexit referendum polling data from Wikipedia:
```{r}
library(rvest)
library(tidyverse)
library(stringr)
url <- "https://en.wikipedia.org/w/index.php?title=Opinion_polling_for_the_United_Kingdom_European_Union_membership_referendum&oldid=896735054"
tab <- read_html(url) %>% html_nodes("table")
polls <- tab[[7]] %>% html_table(fill = TRUE)
```

You will use a variety of string processing techniques learned in this section to reformat these data.


#### Question 5
Some rows in this table do not contain polls. You can identify these by the lack of the percent sign (%) in the Remain column.

Update polls by changing the column names to c("dates", "remain", "leave", "undecided", "lead", "samplesize", "pollster", "poll_type", "notes") and only keeping rows that have a percent sign (%) in the remain column.

How many rows remain in the polls data frame?
```{r}
colnames(polls) <- c("dates", "remain", "leave", "undecided", "lead", "samplesize", "pollster", "poll_type", "notes")
head(polls)
polls <- polls[str_detect(polls$remain,"%"), -9]
nrow(polls)
```
```{r}
polls$undecided
str_replace(polls$undecided, "N/A", "0%")
```

