---
title: 'Data Science: MachineR Basics - HarvardX: PH125.1x'
author: 'Luiz Cunha'
date: '2019-08-19'
output: html_notebook
---

```{r include=FALSE}
# Set knitr options for knitting code into the report:
# - echo=TRUE: print out code (echo), though any results/output would still be displayed
# - cache=TRUE: Save results so that code blocks aren't re-run unless code changes (cache),
# _or_ a relevant earlier code block changed (autodep), but don't re-run if the
# only thing that changed was the comments (cache.comments)
# - message, warning = FALSE: Don't clutter R output with messages or warnings
# This _will_ leave error messages showing up in the knitted report
# - results="hide": hide the results/output (but here the code would still be displayed)
# - include=FALSE:  chunk is evaluated, but neither the code nor its output is displayed
# - eval=FALSE: chunk is not evaluated
knitr::opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE)
```

# Introduction and Welcome

## Introduction and Welcome

Welcome to Data Science: R Basics! We're excited to have you join us in this course, which is designed to teach you the basics of analyzing, visualizing, and working with data using R.

This is the first course in the HarvardX Professional Certificate in Data Science, a series of courses that prepare you to do data analysis in R, from simple computations to machine learning.

This course assumes you are comfortable with basic math, algebra and logical operations. You do not need to have coding experience as you will have ample opportunities in this course to practice your coding skills. HarvardX has partnered with DataCamp for assignments in R that allow you to program directly in a browser-based interface. You will, however, need to download the R software to complete the assessments on the edX platform.

Using a combination of guided introduction through short video lectures and more independent in-depth exploration, you will get to practice your new R skills on real-life applications.

**In this course, you will:**

* Learn to read, extract, and create datasets in R
* Learn to perform a variety of operations on datasets using R
* Learn to write your own functions/sub-routines in R

**Course overview**

* Section 1: R Basics, Functions, Data types
You will get started with R, learn about its functions and data types. 

* Section 2: Vectors, Sorting
You will learn to operate on vectors and advanced functions such as sorting. 

* Section 3: Indexing, Data Manipulation, Plots 
You will learn to wrangle and visualize data.

* Section 4: Programming Basics 
You will learn to use general programming features like ‘if-else’, and ‘for loop’ commands, and write your own functions to perform various operations on datasets.


# Section 1: R Basics, Functions and Data Types

## 1.1 Motivation

### 1.1.1 Overview

Section 1 introduces you to R Basics, Functions, and Datatypes.

In Section 1, you will learn to:

* Appreciate the rationale for data analysis using R.
* Define objects and perform basic arithmetic and logical operations.
* Use pre-defined functions to perform operations on objects.
* Distinguish between various data types.

### 1.1.2 Motivation

**Key points**

* In the first few videos we will go over logistics and introduce general concepts.
* Rather than cover every R skill you need, we will provide the building blocks so you can grow your knowledge as you go through the series.
* Because we better retain knowledge by learning through solving a specific problem, we will use a motivating case study - crime in the United States - to answer specific questions.

### 1.1.3 Getting Started

**Key points**

* R was developed by statisticians and data analysts as an interactive environment for data analysis.
* Some of the advantages of R are that 
  + 1. it is free and open source, 
  + 2. it has the capability to save scripts, 
  + 3. there are numerous resources for learning
  + 4. it is easy for developers to share software implementation.
* Expressions are evaluated in the R console when you type the expression into the console and hit Return.
* A great advantage of R over point and click analysis software is that you can save your work as scripts.
* "Base R" is what you get after you first install R. Additional components are available via packages.

### 1.1.4 Installing R and R Studio
...


## 1.2 R Basics

### 1.2.1 R Basics

**Key Points**

* To define a variable, we may use the assignment symbol “<-“.
* There are two ways to see the value stored in a variable: (1) type the variable into the console and hit Return, or (2) type print(“variable name”) and hit Return.
* Objects are stuff that is stored in R.  They can be variables, functions, etc.
* The function **ls()** shows the names of the objects saved in your workspace.

### 1.2.2 Functions

**Key points**

* In general, to evaluate a function we need to use parentheses. If we type a function without parenthesis, R shows us the code for the function. Most functions also require an argument, that is, something to be written inside the parenthesis.
* To access help files, we may use the help function help(“function name”), or write the question mark followed by the function name.
* The help file shows you the arguments the function is expecting, some of which are required and some are optional. If an argument is optional, a default value is assigned with the equal sign. The args() function also shows the arguments a function needs.
* To specify arguments, we use the equals sign. If no argument name is used, R assumes you’re entering arguments in the order shown in the help file.
* Creating and saving a script makes code much easier to execute.
* To make your code more readable, use intuitive variable names and include comments (using the “#” symbol) to remind yourself why you wrote a particular line of code.


## 1.3 Data Types

### 1.3.1 Data Types

**Key points**

* The function “class” helps us determine the type of an object.
* Data frames can be thought of as tables with rows representing observations and columns representing different variables.
* To access data from columns of a data frame, we use the dollar sign symbol, which is called the accessor.
* A vector is an object consisting of several entries and can be a numeric vector, a character vector, or a logical vector.
* We use quotes to distinguish between variable names and character strings.
* **Factors** are useful for storing categorical data, and are more memory efficient than storing characters.

**Code**
```{r}
# loading the dslabs package and the murders dataset
library(dslabs)
data(murders)

# determining that the murders dataset is of the "data frame" class
class(murders)
# finding out more about the structure of the object
str(murders)
# showing the first 6 lines of the dataset
head(murders)

# using the accessor operator to obtain the population column
murders$population
# displaying the variable names in the murders dataset
names(murders)
# determining how many entries are in a vector
pop <- murders$population
length(pop)
# vectors can be of class numeric and character
class(pop)
class(murders$state)

# logical vectors are either TRUE or FALSE
z <- 3 == 2
z
class(z)

# factors are another type of class
class(murders$region)
# obtaining the levels of a factor
levels(murders$region)
```

