---
title: 'Data Science: Wrangling - HarvardX: PH125.6x'
author: 'Luiz Cunha'
date: '2019-08-13'
output: html_notebook
---

# Final: Comprehensive Assessment

## Comprehensive Assessment: Puerto Rico Hurricane Mortality

### Project Introduction

On September 20, 2017, Hurricane Maria made landfall on Puerto Rico. It was the worst natural disaster on record in Puerto Rico and the deadliest Atlantic hurricane since 2004. However, Puerto Rico's official death statistics only tallied 64 deaths caused directly by the hurricane (due to structural collapse, debris, floods and drownings), an undercount that slowed disaster recovery funding. The majority of the deaths resulted from infrastructure damage that made it difficult to access resources like clean food, water, power, healthcare and communications in the months after the disaster, and although these deaths were due to effects of the hurricane, they were not initially counted.

In order to correct the misconception that few lives were lost in Hurricane Maria, statisticians analyzed how death rates in Puerto Rico changed after the hurricane and estimated the excess number of deaths likely caused by the storm. This analysis suggested that the actual number of deaths in Puerto Rico was 2,975 (95% CI: 2,658-3,290) over the 4 months following the hurricane, much higher than the original count.

We will use your new data wrangling skills to extract actual daily mortality data from Puerto Rico and investigate whether the Hurricane Maria had an immediate effect on daily mortality compared to unaffected days in September 2015-2017.

```{r}
library(tidyverse)
library(pdftools)
options(digits = 3)    # report 3 significant digits
```


### Puerto Rico Hurricane Mortality: Part 1

#### Question 1
In the extdata directory of the dslabs package, you will find a PDF file containing daily mortality data for Puerto Rico from Jan 1, 2015 to May 31, 2018. You can find the file like this:
```{r}
fn <- system.file("extdata", "RD-Mortality-Report_2015-18-180531.pdf", package="dslabs")
```

Find and open the file or open it directly from RStudio. On a Mac, you can type:
```{r}
system2("open", args = fn)
```



#### Question 2
We are going to create a tidy dataset with each row representing one observation. The variables in this dataset will be year, month, day and deaths.  
Use the pdftools package to read in fn using the pdf_text function. Store the results in an object called txt.  
Describe what you see in txt.  

```{r}
txt <- pdf_text(fn)
class(txt)
str(txt)
dim(txt)
```

#### Question 3
Extract the ninth page of the PDF file from the object txt, then use the str_split function from the stringr package so that you have each line in a different entry. The new line character is \n. Call this string vector x.  
Look at x. What best describes what you see?

```{r}
page_9 <- txt[9]
page_9
x <- str_split(page_9, "\n")
class(x)
length(x)
```

#### Question 4
Define s to be the first entry of the x object.  
What kind of object is s?
```{r}
s <- x[[1]]
class(s)
length(s)
s
```

#### Question 5
When inspecting the string we obtained above, we see a common problem: white space before and after the other characters. Trimming is a common first step in string processing. These extra spaces will eventually make splitting the strings hard so we start by removing them.  
We learned about the command str_trim that removes spaces at the start or end of the strings. Use this function to trim s and assign the result to s again.

After trimming, what single character is the last character of element 1 of s?
```{r}
s <- str_trim(s)
s[1] # print string, visually inspect last character
```

#### Question 6
We want to extract the numbers from the strings stored in s. However, there are a lot of non-numeric characters that will get in the way. We can remove these, but before doing this we want to preserve the string with the column header, which includes the month abbreviation.

Use the *str_which* function to find the row with the header. Save this result to *header_index*.  
Hint: find the first string that matches the pattern "2015" using the str_which function.

What is the value of header_index?
```{r}
header_index <- str_which(s, pattern="2015")[1]
header_index
```

#### Question 7
We want to extract two objects from the header row: *month* will store the month and *header* will store the column names.  
Save the content of the header row into an object called header, then use str_split to help define the two objects we need.

What is the value of month?
```{r}
tmp <- str_split(s[header_index], pattern='\\s+', simplify=TRUE)
month <- tmp[1]
header <- tmp[-1]
```


### Puerto Rico Hurricane Mortality: Part 2

#### Question 8
Notice that towards the end of the page defined by s you see a "Total" row followed by rows with other summary statistics. Create an object called tail_index with the index of the "Total" entry.
What is the value of tail_index?
```{r}
tail_index <- str_which(s, pattern="Total")
tail_index
```

#### Question 9
Because our PDF page includes graphs with numbers, some of our rows have just one number (from the y-axis of the plot). Use the str_count function to create an object n with the count of numbers in each row.  
How many rows have a single number in them?
```{r}
n <- str_count(s, pattern='\\d+')
which(n==1)
```

#### Question 10
We are now ready to remove entries from rows that we know we don't need. The entry header_index and everything before it should be removed. Entries for which n is 1 should also be removed, and the entry tail_index and everything that comes after it should be removed as well.

How many entries remain in s?
```{r}
s <- s[-c(1:header_index, which(n==1), tail_index:length(s))]
length(s)
```

#### Question 11
Now we are ready to remove all text that is not a digit or space. Do this using regular expressions (regex) and the str_remove_all function.
In regex, using the ^ inside the square brackets [] means not, like the ! means not in !=. To define the regex pattern to catch all non-numbers, you can type [^\\d]. But remember you also want to keep spaces.

Which of these commands produces the correct output?
```{r}
s <- str_remove_all(s, "[^\\d\\s]")
s
```


#### Question 12
Use the str_split_fixed function to convert s into a data matrix with just the day and death count data:
```{r}
s <- str_split_fixed(s, "\\s+", n = 6)[,1:5]
```

Now you are almost ready to finish. Add column names to the matrix: the first column should be day and the next columns should be the header. Convert all values to numeric. Also, add a column with the month. Call the resulting object tab.
```{r}
# colnames(s) <- c('day', header)
# parse_number(s[,1])
# tab <- data.frame(parse_number(s[,1]), parse_number(s[,2]), parse_number(s[,3]), parse_number(s[,4]), parse_number(s[,5]))
# colnames(tab) <- colnames(s)
# tab <- cbind(tab, month="SEP")
tab <- s %>% 
    as_data_frame() %>% 
    setNames(c("day", header)) %>%
    mutate_all(as.numeric)
tab
```
```{r}
# What was the mean number of deaths per day in September 2015?
mean(tab$`2015`)

# What is the mean number of deaths per day in September 2016?
mean(tab$`2016`)

# Hurricane Maria hit Puerto Rico on September 20, 2017. What was the mean number of deaths per day from September 1-19, 2017, before the hurricane hit?
mean(tab$`2017`[1:19])

# What was the mean number of deaths per day from September 20-30, 2017, after the hurricane hit?
mean(tab$`2017`[20:30])
```


#### Question 13
Finish it up by changing tab to a tidy format, starting from this code outline:
```{r}
tab <- tab %>% gather(year, deaths, -day) %>%
    mutate(deaths = as.numeric(deaths))
tab
```

#### Question 14
Make a plot of deaths versus day with color to denote year. Exclude 2018 since we have no data. Add a vertical line at day 20, the day that Hurricane Maria hit in 2017.
```{r}
tab %>%
    filter(year<2018) %>%
    ggplot(aes(day,deaths, color=year)) + 
    geom_line() +
    geom_vline(xintercept=20)
```

Which of the following are TRUE?

* September 2015 and 2016 deaths by day are roughly equal to each other.
* The day with the most deaths was the day of the hurricane: September 20, 2017.
* After the hurricane in September 2017, there were over 100 deaths per day every day for the rest of the month.
* No days before September 20, 2017 have over 100 deaths per day.
