---
title: "Section 5: Data Visualization Principles"
output: html_notebook
---

# Overview

Section 5 covers some general principles that can serve as guides for effective data visualization.
After completing Section 5, you will:

* understand basic principles of effective data visualization.
* understand the importance of keeping your goal in mind when deciding on a visualization approach.
* understand principles for encoding data, including position, aligned lengths, angles, area, brightness, and color hue.
* know when to include the number zero in visualizations.
* be able to use techniques to ease comparisons, such as using common axes, putting visual cues to be compared adjacent to one another, and using color effectively.


# 5.1 Data Visualization Principles: Part #1


## 5.1.1 Intro to Data Visualization Principles

**Key points**

* We aim to provide some general guidelines for effective data visualization.
* We show examples of plot styles to avoid, discuss how to improve them, and use these examples to explain research-based principles for effective visualization.
* When choosing a visualization approach, keep your goal and audience in mind.


## 5.1.2 Encoding Data Using Visual Cues

**Key points**

* Visual cues for encoding data include position, length, angle, area, brightness and color hue.
* Position and length are the preferred way to display quantities, followed by angles, which are preferred over area. Brightness and color are even harder to quantify but can sometimes be useful.
* Pie charts represent visual cues as both angles and area, while donut charts use only area. Humans are not good at visually quantifying angles and are even worse at quantifying area. Therefore pie and donut charts should be avoided - use a bar plot instead. If you must make a pie chart, include percentages as labels.
* Bar plots represent visual cues as position and length. Humans are good at visually quantifying linear measures, making bar plots a strong alternative to pie or donut charts.


## 5.1.3 Know when to include 0

**Key points**

* When using bar plots, always start at 0. It is deceptive not to start at 0 because bar plots imply length is proportional to the quantity displayed. Cutting off the y-axis can make differences look bigger than they actually are.
* When using position rather than length, it is not necessary to include 0 (scatterplot, dot plot, boxplot).


## 5.1.4 Do not distort quantities

**Key points**

* Make sure your visualizations encode the correct quantities.
* For example, if you are using a plot that relies on circle area, make sure the area (rather than the radius) is proportional to the quantity.


## 5.1.5 Order by a Meaningful Value

**Key points**

* It is easiest to visually extract information from a plot when categories are ordered by a meaningful value. The exact value on which to order will depend on your data and the message you wish to convey with your plot.
* The default ordering for categories is alphabetical if the categories are strings or by factor level if factors. However, we rarely want alphabetical order.



# 5.2 Data Visualization Principles: Part #2


## 5.2.1 Show the Data

**Key points**

* A dynamite plot - a bar graph of group averages with error bars denoting standard errors - provides almost no information about a distribution.
* By showing the data, you provide viewers extra information about distributions.
* Jitter is adding a small random shift to each point in order to minimize the number of overlapping points. To add jitter, use the  geom_jitter geometry instead of geom_point. (See example below.)
* Alpha blending is making points somewhat transparent, helping visualize the density of overlapping points. Add an alpha argument to the geometry.

**Code**

```{r}
# dot plot showing the data
heights %>% ggplot(aes(sex, height)) + geom_point()

# jittered, alpha blended point plot
heights %>% ggplot(aes(sex, height)) + geom_jitter(width = 0.1, alpha = 0.2)
```


## 5.2.2 Ease Comparisons: Use Common Axes

**Key points**

* Ease comparisons by keeping axes the same when comparing data across multiple plots.
* Align plots vertically to see horizontal changes. Align plots horizontally to see vertical changes.
* Bar plots are useful for showing one number but not useful for showing distributions.


## 5.2.3 Consider Transformations

**Key points**

* Use transformations when warranted to ease visual interpretation.
* The log transformation is useful for data with multiplicative changes. The logistic transformation is useful for fold changes in odds. The square root transformation is useful for count data.
* We learned how to apply transformations earlier in the course.


## 5.2.4 Ease Comparisons: Compared Visual Cues Should Be Adjacent

**Key points**

* When two groups are to be compared, it is optimal to place them adjacent in the plot.
* Use color to encode groups to be compared.
* Consider using a color blind friendly palette like the one in this video.

**Code**
```{r}
color_blind_friendly_cols <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
p1 <- data.frame(x = 1:8, y = 1:8, col = as.character(1:8)) %>%
    ggplot(aes(x, y, color = col)) +
    geom_point(size = 5)
p1 + scale_color_manual(values = color_blind_friendly_cols)
```


## 5.2.5 Assessment

### Exercise 1: Customizing plots - watch and learn 0 / 100 XP
To make the plot on the right in the exercise from the last set of assessments, we had to reorder the levels of the states' variables.

**Instructions**

* Redefine the state object so that the levels are re-ordered by rate.
* Print the new object state and its levels so you can see that the vector is now re-ordered by the levels.

**Code**

```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
dat <- us_contagious_diseases %>%
filter(year == 1967 & disease=="Measles" & !is.na(population)) %>% mutate(rate = count / population * 10000 * 52 / weeks_reporting)
state <- dat$state 
rate <- dat$count/(dat$population/10000)*(52/dat$weeks_reporting)
state <- reorder(state, rate)
print(state)
levels(state)
```


### Exercise 2: Customizing plots - redefining 0 / 100 XP

Now we are going to customize this plot a little more by creating a rate variable and reordering by that variable instead.

**Instructions**

* Add a single line of code to the definition of the dat table that uses mutate to reorder the states by the rate variable.
* The sample code provided will then create a bar plot using the newly defined dat.

**Code**

```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
data(us_contagious_diseases)
dat <- us_contagious_diseases %>% filter(year == 1967 & disease=="Measles" & count>0 & !is.na(population)) %>% 
  mutate(rate = count / population * 10000 * 52 / weeks_reporting) %>% 
  mutate(state = reorder(state, rate))
dat %>%
      ggplot(aes(state, rate)) +
      geom_bar(stat="identity") +
      coord_flip()
```

### Exercise 3: Showing the data and customizing plots 0 / 50 XP

Say we are interested in comparing gun homicide rates across regions of the US. We see this plot:

library(dplyr)
library(ggplot2)
library(dslabs)
data("murders")
murders %>% mutate(rate = total/population*100000) %>%
  group_by(region) %>%
  summarize(avg = mean(rate)) %>%
  mutate(region = factor(region)) %>%
  ggplot(aes(region, avg)) +
  geom_bar(stat="identity") +
  ylab("Murder Rate Average")
and decide to move to a state in the western region. What is the main problem with this interpretation?

### Exercise 4: Making a box plot

To further investigate whether moving to the western region is a wise decision, let's make a box plot of murder rates by region, showing all points.

**Instructions**

* Make a box plot of the murder rates by region.
* Order the regions by their median murder rate.
* Show all of the points on the box plot.

**Code**
```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
data("murders")
murders %>% 
  mutate(rate = total/population*100000) %>%
  mutate(region = reorder(region,rate, median)) %>%
  ggplot(aes(x=region, y=rate)) +
  geom_boxplot() +
  geom_point()
```