---
title: 'Data Science: MachineR Basics - HarvardX: PH125.1x'
author: 'Luiz Cunha'
date: '2019-08-19'
output: html_notebook
---

```{r include=FALSE}
# Set knitr options for knitting code into the report:
# - echo=TRUE: print out code (echo), though any results/output would still be displayed
# - cache=TRUE: Save results so that code blocks aren't re-run unless code changes (cache),
# _or_ a relevant earlier code block changed (autodep), but don't re-run if the
# only thing that changed was the comments (cache.comments)
# - message, warning = FALSE: Don't clutter R output with messages or warnings
# This _will_ leave error messages showing up in the knitted report
# - results="hide": hide the results/output (but here the code would still be displayed)
# - include=FALSE:  chunk is evaluated, but neither the code nor its output is displayed
# - eval=FALSE: chunk is not evaluated
knitr::opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE)
```

# Section 2: Vectors, Sorting

## Overview

Section 2 introduces you to vectors and functions such as sorting.

In Section 2.1, you will: 

* Create numeric and character vectors. 
* Name the columns of a vector. 
* Generate numeric sequences. 
* Access specific elements or parts of a vector.
* Coerce data into different data types as needed. 

In Section 2.2, you will: 

* Sort vectors in ascending and descending order. 
* Extract the indices of the sorted elements from the original vector. 
* Find the maximum and minimum elements, as well as their indices, in a vector. 
* Rank the elements of a vector in increasing order. 

In Section 2.3, you will: 

* Perform arithmetic between a vector and a single number.
* Perform arithmetic between two vectors of the same length. 


## 2.1 Vectors

### 2.1.1 Vectors

**Key Points**

* The function **c()**, which stands for concatenate, is useful for creating vectors.
* Another useful function for creating vectors is the **seq()** function, which generates sequences.
** Subsetting lets us access specific parts of a vector by using square brackets to access elements of a vector.

**Code**
```{r}
# We may create vectors of class numeric or character with the concatenate function
codes <- c(380, 124, 818)
country <- c("italy", "canada", "egypt")

# We can also name the elements of a numeric vector
# Note that the two lines of code below have the same result
codes <- c(italy = 380, canada = 124, egypt = 818)
codes <- c("italy" = 380, "canada" = 124, "egypt" = 818)

# We can also name the elements of a numeric vector using the names() function
codes <- c(380, 124, 818)
country <- c("italy","canada","egypt")
names(codes) <- country

# Using square brackets is useful for subsetting to access specific elements of a vector
codes[2]
codes[c(1,3)]
codes[1:2]

# If the entries of a vector are named, they may be accessed by referring to their name
codes["canada"]
codes[c("egypt","italy")]
```

### 2.1.2 Vector Coercion

**Key points**

* In general, **coercion** is an attempt by R to be flexible with data types by guessing what was meant when an entry does not match the expected. For example, when defining x as
    >x <- c(1, "canada", 3)

    R coerced the data into characters. It guessed that because you put a character string in the vector, you meant the 1 and 3 to actually be character strings "1" and “3”.

* The function **as.character()** turns numbers into characters.
* The function **as.numeric()** turns characters into numbers.
* In R, missing data is assigned the value **NA**.


## 2.2 Sorting

### 2.2.1 Sorting

**Key points**

* The function **sort()** sorts a vector in increasing order.
* The function **order()** produces the indices needed to obtain the sorted vector, e.g. a result of  2 3 1 5 4 means the sorted vector will be produced by listing the 2nd, 3rd, 1st, 5th, and then 4th item of the original vector.
* The function **rank()** gives us the ranks of the items in the original vector.
* The function **max()** returns the largest value while **which.max()** returns the index of the largest value. The functions **min()** and **which.min()** work similarly for minimum values.


## 2.3 Vector Arithmetic

### 2.3.1 Vector Arithmetic

**Key points**

* In R, **arithmetic operations on vectors occur element-wise**.

**Code**
```{r}
# The name of the state with the maximum population is found by doing the following
murders$state[which.max(murders$population)]

# how to obtain the murder rate
murder_rate <- murders$total / murders$population * 100000

# ordering the states by murder rate, in decreasing order
murders$state[order(murder_rate, decreasing=TRUE)]
```


