---
title: "Section 1: Discrete Probability"
output: html_notebook
---

# 1.2 Combinations and Permutations


## 1.2.1 Combinations and Permutations

**ORDER and REPLACEMENT**

* Cardinal:     with ORDER && with REPLACEMENT
* Permutations: with ORDER && w/o  REPLACEMENT 
* Combinations: w/o  ORDER && w/o  REPLACEMENT

**Key points**

* paste joins two strings and inserts a space in between.
* expand.grid gives the combinations of 2 vectors or lists.
* permutations(n,r) from the gtools package lists the different ways that r items can be selected from a set of n options when order matters.
* combinations(n,r) from the gtools package lists the different ways that r items can be selected from a set of n options when order does not matter.

**Code: Introducing paste and expand.grid**
```{r}
# joining strings with paste (nb: ~zip in Python)
number <- "Three"
suit <- "Hearts"
paste(number, suit)

# joining vectors element-wise with paste
paste(letters[1:5], as.character(1:5))

# generating combinations of 2 vectors with expand.grid
expand.grid(pants = c("blue", "black"), shirt = c("white", "grey", "plaid"))
```

**Code: Generating a deck of cards**
```{r}
suits <- c("Diamonds", "Clubs", "Hearts", "Spades")
numbers <- c("Ace", "Deuce", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nine", "Ten", "Jack", "Queen", "King")
deck <- expand.grid(number = numbers, suit = suits)
deck <- paste(deck$number, deck$suit)

# probability of drawing a king
kings <- paste("King", suits)
mean(deck %in% kings)
```

**Code: Permutations and combinations**
```{r}
library(gtools)
#permutations(5,2)    # ways to choose 2 numbers in order from 1:5

all_phone_numbers <- permutations(10, 7, v = 0:9)
n <- nrow(all_phone_numbers)
index <- sample(n, 5)
all_phone_numbers[index,]

permutations(3,2)    # order matters
combinations(3,2)    # order does not matter
```

**Code: Probability of drawing a second king given that one king is drawn**
```{r}
hands <- permutations(52,2, v = deck)
first_card <- hands[,1]
second_card <- hands[,2]
sum(first_card %in% kings)

sum(first_card %in% kings & second_card %in% kings) / sum(first_card %in% kings)
```

**Code: Probability of a natural 21 in blackjack**
```{r}
aces <- paste("Ace", suits)

facecard <- c("King", "Queen", "Jack", "Ten")
facecard <- expand.grid(number = facecard, suit = suits)
facecard <- paste(facecard$number, facecard$suit)

hands <- combinations(52, 2, v=deck) # all possible hands

# probability of a natural 21 given that the ace is listed first in `combinations`
mean(hands[,1] %in% aces & hands[,2] %in% facecard)

# probability of a natural 21 checking for both ace first and ace second
mean((hands[,1] %in% aces & hands[,2] %in% facecard)|(hands[,2] %in% aces & hands[,1] %in% facecard))
```

**Code: Monte Carlo simulation of natural 21 in blackjack**  
Note that your exact values will differ because the process is random and the seed is not set.
```{r}
# code for one hand of blackjack
hand <- sample(deck, 2)
hand

# code for B=10,000 hands of blackjack
B <- 10000
results <- replicate(B, {
  hand <- sample(deck, 2)
  (hand[1] %in% aces & hand[2] %in% facecard) | (hand[2] %in% aces & hand[1] %in% facecard)
})
mean(results)
```


## 1.2.2 The Birthday Problem

**Key points**

* *duplicated* takes a vector and returns a vector of the same length with TRUE for any elements that have appeared previously in that vector.
* We can compute the probability of shared birthdays in a group of people by modeling birthdays as random draws from the numbers 1 through 365. We can then use this sampling model of birthdays to run a Monte Carlo simulation to estimate the probability of shared birthdays.

**Code: The birthday problem**
```{r}
# checking for duplicated bdays in one 50 person group
n <- 50
bdays <- sample(1:365, n, replace = TRUE)    # generate n random birthdays
any(duplicated(bdays))    # check if any birthdays are duplicated

# Monte Carlo simulation with B=10000 replicates
B <- 10000
results <- replicate(B, {    # returns vector of B logical values
    bdays <- sample(1:365, n, replace = TRUE)
    any(duplicated(bdays))
})
mean(results)    # calculates proportion of groups with duplicated bdays
```


# 1.2.3 Function sapply 

**Key points**

* Some functions automatically apply element-wise to vectors, such as sqrt and *.
* However, other functions do not operate element-wise by default. This includes functions we define ourselves.
* The function sapply(x, f) allows any other function f to be applied element-wise to the vector x.
* The probability of an event happening is 1 minus the probability of that event not happening:  
$$Pr(event)=1−Pr(no event)$$
* We can compute the probability of shared birthdays mathematically:  
$$Pr(shared birthdays) = 1-Pr(no shared birthdays) = 1-(1×364/365×363/365×...×(365-n+1)/365$$

**Code: Function for calculating birthday problem Monte Carlo simulations for any value of n**

Note that the function body of *compute_prob* is the code that we wrote in the previous video. If we write this code as a function, we can use *sapply* to apply this function to several values of n.

```{r}
# function to calculate probability of shared bdays across n people
compute_prob <- function(n, B = 10000) {
    same_day <- replicate(B, {
        bdays <- sample(1:365, n, replace = TRUE)
        any(duplicated(bdays))
    })
    mean(same_day)
}

n <- seq(1, 60)
```

*Code: Element-wise operation over vectors and sapply*
```{r}
x <- 1:10
sqrt(x)    # sqrt operates on each element of the vector

y <- 1:10
x*y    # * operates element-wise on both vectors

compute_prob(n)    # does not iterate over the vector n without sapply

x <- 1:10
sapply(x, sqrt)    # this is equivalent to sqrt(x)

prob <- sapply(n, compute_prob)    # element-wise application of compute_prob to n
plot(n, prob)
```

*Code: Computing birthday problem probabilities with sapply*
```{r}
# function for computing exact probability of shared birthdays for any n
exact_prob <- function(n){
    prob_unique <- seq(365, 365-n+1)/365   # vector of fractions for mult. rule
    1 - prod(prob_unique)    # calculate prob of no shared birthdays and subtract from 1
}

# applying function element-wise to vector of n values
eprob <- sapply(n, exact_prob)

# plotting Monte Carlo results and exact probabilities on same graph
plot(n, prob)    # plot Monte Carlo results
lines(n, eprob, col = 'red')    # add line for exact prob
```


## 1.2.4 How Many Monte Carlo Experiments are enough?

**Key points**

* The larger the number of Monte Carlo replicates *B* , the more accurate the estimate.
* Determining the appropriate size for *B* can require advanced statistics.
* One practical approach is to try many sizes for *B* and look for sizes that provide stable estimates.

**Code: Estimating a practical value of B**

This code runs Monte Carlo simulations to estimate the probability of shared birthdays using several *B* values and plots the results. When *B* is large enough that the estimated probability stays stable, then we have selected a useful value of *B*. 
```{r}
B <- 10^seq(1, 5, len = 100)    # defines vector of many B values
compute_prob <- function(B, n = 22){   # function to run Monte Carlo simulation with each B
    same_day <- replicate(B, {
        bdays <- sample(1:365, n, replace = TRUE)
        any(duplicated(bdays))
    })
    mean(same_day)
}
prob <- sapply(B, compute_prob)    # apply compute_prob to many values of B
plot(log10(B), prob, type = "l")    # plot a line graph of estimates 
```

## 1.2.5 Assessment: Monte Carlo simulation for Celtics winning a game

```{r}
B <- 10
celtic_wins <- replicate(B, {
  simulated_games <- sample(c("lose","win"), 4, replace = TRUE, prob = c(0.6, 0.4))
  any(simulated_games=='win')
})
mean(celtic_wins)
```