---
title: "Section 1: Discrete Probability"
output: html_notebook
---

# 1.3 Addition Rule and Monty Hall


## 1.3.1 The Addition Rule

**!!! ORDER and REPLACEMENT**

* Cardinal:     with ORDER && with REPLACEMENT
* Permutations: with ORDER && w/o  REPLACEMENT 
* Combinations: w/o  ORDER && w/o  REPLACEMENT

**Key points**

* The addition rule states that the probability of event *A* or event *B* happening is the probability of event *A* plus the probability of event *B* minus the probability of both events *A* and *B* happening together.
$$Pr(A\ or\ B) = Pr(A) + Pr(B) - Pr(A\ and\ B)$$
* Note that *(A or B)* is equivalent to *(A|B)*.

**Example: The addition rule for a natural 21 in blackjack**

We apply the addition rule where  𝐴  = drawing an ace then a facecard and  𝐵  = drawing a facecard then an ace. Note that in this case, both events A and B cannot happen at the same time, so  Pr(A and B)=0 .


## 1.3.2 The Monty Hall Problem

**Key points**

* Monte Carlo simulations can be used to simulate random outcomes, which makes them useful when exploring ambiguous or less intuitive problems like the Monty Hall problem.
* In the Monty Hall problem, contestants choose one of three doors that may contain a prize. Then, one of the doors that was not chosen by the contestant and does not contain a prize is revealed. The contestant can then choose whether to stick with the original choice or switch to the remaining unopened door.
* Although it may seem intuitively like the contestant has a 1 in 2 chance of winning regardless of whether they stick or switch, Monte Carlo simulations demonstrate that the actual probability of winning is 1 in 3 with the stick strategy and 2 in 3 with the switch strategy.

**Code: Monte Carlo simulation of stick strategy**

```{r}
B <- 10000
stick <- replicate(B, {
  doors <- as.character(1:3)
  prize <- sample(c("car","goat","goat"))    # puts prizes in random order
  prize_door <- doors[prize == "car"]    # note which door has prize
  my_pick  <- sample(doors, 1)    # note which door is chosen
  show <- sample(doors[!doors %in% c(my_pick, prize_door)],1)    # open door with no prize that isn't chosen
  stick <- my_pick    # stick with original door
  stick == prize_door    # test whether the original door has the prize
})
mean(stick)    # probability of choosing prize door when sticking
```

**Code: Monte Carlo simulation of switch strategy**
```{r}
switch <- replicate(B, {
  doors <- as.character(1:3)
  prize <- sample(c("car","goat","goat"))    # puts prizes in random order
  prize_door <- doors[prize == "car"]    # note which door has prize
  my_pick  <- sample(doors, 1)    # note which door is chosen first
  show <- sample(doors[!doors %in% c(my_pick, prize_door)], 1)    # open door with no prize that isn't chosen
  switch <- doors[!doors%in%c(my_pick, show)]    # switch to the door that wasn't chosen first or opened
  switch == prize_door    # test whether the switched door has the prize
})
mean(switch)    # probability of choosing prize door when switching
```

## 1.3.3 Assessment


### Exercise 1. The Cavs and the Warriors

Two teams, say the Cavs and the Warriors, are playing a seven game championship series. The first to win four games wins the series. The teams are equally good, so they each have a 50-50 chance of winning each game.

If the Cavs lose the first game, what is the probability that they win the series?

**Instructions**

* Assign the number of remaining games to the variable *n*.
* Assign a variable *outcomes* as a vector of possible outcomes in a single game, where 0 indicates a loss and 1 indicates a win for the Cavs.
* Assign a variable *l* to a list of all possible outcomes in all remaining games. Use the *rep* function to create a list of n games, where each game consists of *list(outcomes)*.
* Use the *expand.grid* function to create a data frame containing all the combinations of possible outcomes of the remaining games.
* Use the *rowSums* function to identify which combinations of game outcomes result in the Cavs winning the number of games necessary to win the series.
* Use the *mean* function to calculate the proportion of outcomes that result in the Cavs winning the series and print your answer to the console.

**Code**

```{r}
# Assign a variable 'n' as the number of remaining games.
n <- 6

# Assign a variable `outcomes` as a vector of possible game outcomes, where 0 indicates a loss and 1 indicates a win for the Cavs.
outcomes <- c(loss=0,win=1)

# Assign a variable `l` to a list of all possible outcomes in all remaining games. Use the `rep` function on `list(outcomes)` to create list of length `n`.
l <- rep(list(outcomes), n)

# Create a data frame named 'possibilities' that contains all combinations of possible outcomes for the remaining games.
possibilities <- expand.grid(l)

# Create a vector named 'results' that indicates whether each row in the data frame 'possibilities' contains enough wins for the Cavs to win the series.
results <- rowSums(possibilities) >=4

# Calculate the proportion of 'results' in which the Cavs win the series. Print the outcome to the console.
mean(results)
```


### Exercise 2. The Cavs and the Warriors - Monte Carlo

Confirm the results of the previous question with a Monte Carlo simulation to estimate the probability of the Cavs winning the series after losing the first game.

**Instructions**

* Use the *replicate* function to replicate the sample code for *B <- 10000* simulations.
* Use the *sample* function to simulate a series of 6 games with random, independent outcomes of either a loss for the Cavs (0) or a win for the Cavs (1) in that order.
* Use the *sum* function to determine whether a simulated series contained at least 4 wins for the Cavs.
* Use the *mean* function to find the proportion of simulations in which the Cavs win at least 4 of the remaining games. Print your answer to the console.

**Code**
```{r}
# The variable `B` specifies the number of times we want the simulation to run. Let's run the Monte Carlo simulation 10,000 times.
B <- 10000

# Use the `set.seed` function to make sure your answer matches the expected result after random sampling.
set.seed(1)

# Create an object called `results` that replicates for `B` iterations a simulated series and determines whether that series contains at least four wins for the Cavs.
results <- replicate(B, {
    series <- sample(c(0,1), 6, replace=TRUE)
    sum(series) >= 4
    })

# Calculate the frequency out of `B` iterations that the Cavs won at least four games in the remainder of the series. Print your answer to the console.
mean(results)
```


### Exercise 3. A and B play a series - part 1

Two teams, A and B, are playing a seven series game series. Team A is better than team B and has a p>0.5 chance of winning each game.

**Instructions**

* Use the function *sapply* to compute the probability, call it *Pr* of winning for *p <- seq(0.5, 0.95, 0.025)*.
* Then plot the result *plot(p, Pr)*.

**Code**
```{r}
# Let's assign the variable 'p' as the vector of probabilities that team A will win.
p <- seq(0.5, 0.95, 0.025)

# Given a value 'p', the probability of winning the series for the underdog team B can be computed with the following function based on a Monte Carlo simulation:
prob_win <- function(p){
  B <- 10000
  result <- replicate(B, {
    b_win <- sample(c(1,0), 7, replace = TRUE, prob = c(1-p, p))
    sum(b_win)>=4
    })
  mean(result)
}

# Apply the 'prob_win' function across the vector of probabilities that team A will win to determine the probability that team B will win. Call this object 'Pr'.
Pr <- sapply(p, prob_win)

# Plot the probability 'p' on the x-axis and 'Pr' on the y-axis.
plot(p, Pr)
```

### Exercise 4. A and B play a series - part 2

Repeat the previous exercise, but now keep the probability that team A wins fixed at *p <- 0.75* and compute the probability for different series lengths. For example, wins in best of 1 game, 3 games, 5 games, and so on through a series that lasts 25 games.

**Instructions**

* Use the *seq* function to generate a list of odd numbers ranging from 1 to 25.
* Use the function *sapply* to compute the probability, call it *Pr*, of winning during series of different lengths.
* Then plot the result *plot(N, Pr)*.

**Code**
```{r}
# Given a value 'p', the probability of winning the series for the underdog team B can be computed with the following function based on a Monte Carlo simulation:
prob_win <- function(N, p=0.75){
      B <- 10000
      result <- replicate(B, {
        b_win <- sample(c(1,0), N, replace = TRUE, prob = c(1-p, p))
        sum(b_win)>=(N+1)/2
        })
      mean(result)
    }

# Assign the variable 'N' as the vector of series lengths. Use only odd numbers ranging from 1 to 25 games.
N <- seq(1,25,2)

# Apply the 'prob_win' function across the vector of series lengths to determine the probability that team B will win. Call this object 'Pr'.
Pr <- sapply(N, prob_win)

# Plot the number of games in the series 'N' on the x-axis and 'Pr' on the y-axis.
plot(N, Pr)
```