---
title: 'Data Science: MachineR Basics - HarvardX: PH125.1x'
author: 'Luiz Cunha'
date: '2019-08-19'
output: html_notebook
---

```{r include=FALSE}
# Set knitr options for knitting code into the report:
# - echo=TRUE: print out code (echo), though any results/output would still be displayed
# - cache=TRUE: Save results so that code blocks aren't re-run unless code changes (cache),
# _or_ a relevant earlier code block changed (autodep), but don't re-run if the
# only thing that changed was the comments (cache.comments)
# - message, warning = FALSE: Don't clutter R output with messages or warnings
# This _will_ leave error messages showing up in the knitted report
# - results="hide": hide the results/output (but here the code would still be displayed)
# - include=FALSE:  chunk is evaluated, but neither the code nor its output is displayed
# - eval=FALSE: chunk is not evaluated
knitr::opts_chunk$set(echo=TRUE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE)
```

# Section 3: Indexing, Data Wrangling, Plots

## Overview

Section 3 introduces the R commands and techniques that help you wrangle, analyze, and visualize data.

In Section 3.1, you will:

* Subset a vector based on properties of another vector. 
* Use multiple logical operators to index vectors.
* Extract the indices of vector elements satisfying one or more logical conditions. 
* Extract the indices of vector elements matching with another vector.
* Determine which elements in one vector are present in another vector. 

In Section 3.2, you will:

* Wrangle data tables using functions in the ‘dplyr’ package. 
* Modify a data table by adding or changing columns. 
* Subset rows in a data table.
* Subset columns in a data table. 
* Perform a series of operations using the pipe operator. 
* Create data frames. 

In Section 3.3, you will:

* Plot data in scatter plots, box plots, and histograms. 


## 3.1 Indexing

### 3.1.1 Indexing

**Key points**

* We can use logicals to index vectors.
* Using the function >sum()on a logical vector returns the number of entries that are true.
* The logical operator “&” makes two logicals true only when they are both true.

### 3.1.2 Indexing Functions

**Key points**

* The function **which()** gives us the entries of a logical vector that are true.
* The function **match()** looks for entries in a vector and returns the index needed to access them.
* We use the function **%in%** if we want to know whether or not each element of a first vector is in a second vector.
Code

```{r}
# to determine the murder rate in Massachusetts we may do the following
ind <- which(murders$state == "Massachusetts")
murder_rate[ind]

# to obtain the indices and subsequent murder rates of New York, Florida, Texas, we do:
ind <- match(c("New York", "Florida", "Texas"), murders$state)
ind
murder_rate[ind]

# to see if Boston, Dakota, and Washington are states
c("Boston", "Dakota", "Washington") %in% murders$state
```


## 3.2 Basic Data Wrangling

### 3.2.1 Basic Data Wrangling

**Key points**

* To change a data table by adding a new column, or changing an existing one, we use the **mutate** function.
* To filter the data by subsetting rows, we use the function filter.
To subset the data by selecting specific columns, we use the select function.
We can perform a series of operations by sending the results of one function to another function using what is called the pipe operator, %>%.

**Code**
```{r}
# installing and loading the dplyr package
library(dplyr)
library(dslabs)
data("murders")

murders <- mutate(murders, rate = total / population * 100000)

# subsetting with filter
filter(murders, rate <= 0.71)

# selecting columns with select
new_table <- select(murders, state, region, rate)

# using the pipe
murders %>% 
    select(state, region, rate) 
%>% filter(rate <= 0.71)
```


### 3.2.2 Creating Data Frames

**Key Points**

* We can use the **data.frame()** function to create data frames.
* By default, the **data.frame()** function turns characters into factors.  To avoid this, we utilize the **stringsAsFactors** argument and set it equal to false.

**Code**
```{r}
# creating a data frame with stringAsFactors = FALSE
grades <- data.frame(names = c("John", "Juan", "Jean", "Yao"), 
                     exam_1 = c(95, 80, 90, 85), 
                     exam_2 = c(90, 85, 85, 90),
                     stringsAsFactors = FALSE)
```


## 3.3 Plots

## 3.3.1 Plots

**Key Points**

* We can create a simple scatterplot using the function **plot()**.
* Histograms are graphical summaries that give you a general overview of the types of values you have.  In R, they can be produced using the **hist()** function.
* **Boxplots** provide a more compact summary of a distribution than a histogram and are more useful for comparing distributions. They can be produced using the boxplot() function.

**Code**
```{r}
# a simple scatterplot of total murders versus population
x <- murders$population / 10^6
y <- murders$total
plot(x, y)

# a histogram of murder rates
hist(murders$rate)

# boxplots of murder rates by region
boxplot(rate~region, data = murders)
```

